---
title: "Hippocampal-to-Ventricle Ratio as a Head-Size-Independent Biomarker: Sex Differences and Cognitive Associations in 27,680 UK Biobank Participants"
running-title: "HVR Sex Differences in UK Biobank"
keywords:
  - Hippocampus
  - Lateral Ventricles
  - Sex Characteristics
  - Magnetic Resonance Imaging
  - Cognitive Aging
  - Biomarkers
  - Reference Values
key-points:
  - Hippocampal sex differences reverse direction depending on head-size adjustment method, spanning 1.5 SD across analytical choices
  - The hippocampal-to-ventricle ratio (HVR) provides consistent sex difference estimates without requiring head-size correction and predicts cognition comparably to adjusted hippocampal volume
  - Males show steeper cross-sectional HVR decline with age; sex-specific normative centile curves are provided for clinical application
author:
  - name: Sofia Fernandez-Lozano
    affiliations:
      - ref: bic
      - ref: ipn
  - name: D. Louis Collins
    affiliations:
      - ref: bic
      - ref: ipn
      - ref: bbme
    email: louis.collins@mcgill.ca
    corresponding: true
affiliations:
  - id: bic
    name: McConnell Brain Imaging Centre
    department: Montreal Neurological Institute
    city: Montreal
    state: Quebec
    country: Canada
  - id: ipn
    name: McGill University
    department: Department of Neurology and Neurosurgery
    city: Montreal
    state: Quebec
    country: Canada
  - id: bbme
    name: McGill University
    department: Department of Biomedical Engineering
    city: Montreal
    state: Quebec
    country: Canada
date: today
abstract: |
  Hippocampal volume is a key biomarker for Alzheimer's disease and brain aging, yet reported sex differences vary substantially depending on head-size adjustment methods. We examined whether the hippocampal-to-ventricle ratio (HVR), a self-normalizing measure, provides more consistent sex difference estimates than conventional adjustment approaches. Using a subset of UK Biobank structural MRI data (N = 27,680; 56% female; ages 45--82), we compared sex differences across four adjustment methods (unadjusted, proportions, stereotaxic, residualized) and in samples matched on age and on intracranial volume (ICV). Hippocampal sex differences reversed direction across methods, ranging from $d = -0.89$ (males larger, unadjusted) to $d = 0.58$ (females larger, proportions), with a range of 1.5 standard deviations across analytical choices. In contrast, HVR showed a consistent female advantage ($d = 0.52$) that persisted in ICV-matched samples ($d = 0.19$), confirming this effect is not a head-size artifact. Males exhibited steeper cross-sectional age-related HVR decline ($1.7\times$ female rate, $p < 10^{-50}$), consistent with earlier-onset male ventricular expansion. Structural equation modeling revealed that HVR predicted general cognitive ability comparably to hippocampal volume ($\beta = 0.04$, overlapping CIs), and unlike residualized hippocampal volume, HVR maintained significant brain-cognition associations in both sexes. We provide sex-specific normative centile curves for clinical application. These findings indicate that apparent hippocampal sex differences largely reflect methodological choices rather than biology, while HVR captures consistent morphological variation related to brain aging that may have clinical utility.
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    include-in-header:
      text: |
        <style>
        .gt_table, .gt_table th, .gt_table td {
          font-size: 14px !important;
        }
        table, th, td {
          font-size: 14px !important;
        }
        </style>
  pdf:
    toc: false
    number-sections: true
    documentclass: article
    geometry:
      - margin=1in
    latex-max-runs: 20
    pdf-engine: xelatex
    keep-tex: true
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage{array}
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{amsmath}
        \usepackage{amssymb}
        \usepackage{textcomp}
        \usepackage{fontspec}
        \setmainfont{Latin Modern Roman}
        \usepackage{newunicodechar}
        \newunicodechar{ε}{\ensuremath{\varepsilon}}
        \usepackage{authblk}
        \renewcommand\Authfont{\normalsize}
        \renewcommand\Affilfont{\small\itshape}
        % Force tables to appear exactly where placed (prevent float drift)
        \usepackage{tikz}
        \usetikzlibrary{shapes.geometric, arrows.meta, positioning}
        \usepackage{float}
        \floatplacement{table}{H}
        % Center longtables
        \setlength{\LTleft}{\fill}
        \setlength{\LTright}{\fill}
  rtf:
    toc: false
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
  eval: true
bibliography: references.bib
csl: human-brain-mapping.csl
---

```{r}
#| label: setup
#| include: false

# Activate renv (quarto runs R with --vanilla, skipping .Rprofile)
renv::load(file.path(getwd(), ".."))

library(here)
library(data.table)
library(gt)
here::i_am("reports-src/manuscript.qmd")
source(here("reports-src/_common.R"))

# Load pre-computed manuscript environment (from R/12_manuscript_objects.R)
ms <- read_rds_safe(here("data/derivatives/manuscript_env.rds"), "manuscript environment")
list2env(ms$scalars, envir = environment())
list2env(ms$data, envir = environment())
list2env(ms$tables, envir = environment())

```

# Introduction

The hippocampus (HC) is a brain structure used as an early biomarker for Alzheimer's disease (AD) and brain aging [@jack_hypothetical_2010; @raz_regional_2005; @jack_prediction_1999]. Sex differences in brain structures have been documented, but their interpretation is controversial [@ritchie_sex_2018; @ruigrok_meta-analysis_2014], largely because findings depend on whether and how head-size adjustments are applied [@sanchis-segura_effects_2020; @nordenskjold_intracranial_2015]. Males have approximately 12% larger intracranial volume (ICV) than females [@ruigrok_meta-analysis_2014], and after head-size correction, sex classification from brain features drops from >80% to approximately 60% [@sanchis-segura_effects_2020]. This methodological sensitivity suggests some reported sex differences may be artifacts of adjustment methods.

Hippocampal atrophy rates predict conversion from mild cognitive impairment to dementia [@rajagopal_estimating_2024], making accurate volume assessment clinically consequential. However, males and females show different rates of age-related brain volume change, with steeper subcortical volume declines in males across multiple structures [@ritchie_sex_2018]. These sex-differential trajectories mean that normative reference values must be sex-stratified to avoid misclassifying normal variation as pathological; recent large-scale brain charting efforts have adopted sex-specific centile modeling for this reason [@bethlehem_brain_2022; @ge_normative_2024].

Multiple adjustment methods exist, each with different assumptions: proportions (V/ICV) assumes perfect allometric scaling, residualization (V $-$ $\beta \times$ ICV) removes linear ICV association, and stereotaxic normalization applies spatial transformation to a template. These methods can yield contradictory conclusions about the same brain regions [@wang_comparison_2024]. For hippocampus, reported sex differences range from males larger (unadjusted; @ruigrok_meta-analysis_2014) to females larger (proportions; @sanchis-segura_effects_2020) to no difference (residualized; @obrien_statistical_2011). The proportions method can produce biologically implausible results, while residualization yields more consistent estimates [@wang_comparison_2024]. No consensus exists on the correct method, and conclusions depend on analytical choices.

The hippocampal-to-ventricle ratio (HVR), defined as HC / (HC + LV) where HC denotes total hippocampal volume and LV denotes the volume of the temporal horns of the lateral ventricles, offers a potential solution. Both structures scale with head size, making their ratio self-normalizing. HVR captures the relationship between hippocampal and ventricular volumes, which show opposite patterns with age—smaller hippocampus and larger ventricles. Schoemaker and colleagues introduced HVR and showed that it provides stronger associations with age and memory than standard hippocampal volume in healthy aging [@schoemaker_hippocampal-ventricle_2019]. Subsequent work shows HVR provides higher effect sizes for distinguishing AD from controls compared to HC alone [@fernandez-lozano_enhanced_2025]. However, properties of HVR in healthy populations and across adjustment methods remain unexplored.

Age-related ventricular expansion reflects concurrent loss of surrounding gray and white matter tissue [@madsen_mapping_2015], and this expansion occurs earlier in males, with abnormal enlargement appearing in the seventies and preferentially affecting the frontal horns [@curra_ventricular_2019]. These sex-differential ventricular trajectories suggest that a ratio incorporating both hippocampal and ventricular volumes may capture sex differences in brain aging that individual volumes miss. Furthermore, brain structure–cognition associations themselves may differ by sex: grey matter volumes predict cognitive profiles with different strength and spatial specificity in older males versus females [@jockwitz_differential_2024], and hippocampal volume predicts associative memory in older women but not men [@zheng_sex_2017]. Whether HVR shows similar sex-differential cognitive associations remains unknown.

We addressed three questions: (1) Do hippocampal sex differences vary across head-size adjustment methods while HVR remains consistent? (2) Does HVR show comparable cognitive associations to conventionally adjusted hippocampal volumes? (3) Do cross-sectional age patterns differ by sex, and do these patterns differ between HC and HVR depending on normalization method?

---

# Materials and Methods

## Data Source and Participants

Data were obtained from UK Biobank (Application 45551) [@sudlow_uk_2015; @miller_multimodal_2016]. The initial imaging sample comprised approximately `r format(n_initial_imaging, big.mark=",")` participants with T1-weighted structural MRI. After exclusions (detailed below), the final analytic sample included N = `r format(n_primary, big.mark=",")` participants aged `r age_min` to `r age_max` years. With this sample size, we had >99% power to detect small effects (d = 0.10) and >80% power to detect very small effects (d = 0.05) for two-group comparisons at $\alpha$ = .05.

Three additional subsamples were defined for specific analyses (@fig-prisma; @tbl-demographics). The **England subsample** (N = `r format(n_england, big.mark=",")`) comprised participants assessed at UK Biobank centres in England and was used for SEM analyses requiring the Index of Multiple Deprivation (IMDP), which is not comparable across UK countries. This subsample provides adequate power for detecting standardized path coefficients $\geq$ 0.02. The **ICV-matched subsample** (N = `r format(n_matched_pairs * 2, big.mark=",")`) comprised female--male pairs matched on age ($\pm 1$ year) and ICV ($\pm 25$ cc) drawn from the primary sample, and was used to isolate sex effects from head-size confounding (Analysis 2). The **longitudinal subsample** (N = 2,935) comprised participants from the primary sample who also had a follow-up imaging session (ses-3), with a mean intersession interval of 2.7 years (SD = 1.1; range: 1.5--5.0). This subsample was used as a sensitivity check for normative model calibration (Analysis 4).

### Exclusion Criteria

Participants were excluded if they had any of the following ICD-10 diagnoses recorded in linked hospital records: F00-F99 (mental and behavioral disorders including dementia, mood disorders, schizophrenia, and anxiety disorders, which are associated with hippocampal volume alterations [@schmaal_subcortical_2016; @cao_volumes_2023; @brosch_reduced_2022]); G00-G99 (neurological diseases including Parkinson's disease, multiple sclerosis, epilepsy, and cerebrovascular disease, which directly affect brain structure [@van_erp_subcortical_2016; @xu_longitudinal_2020; @koenig_hippocampal_2014; @tai_detection_2025]); and Q00-Q07 (congenital CNS malformations affecting normal brain development). This exclusion strategy selected a sample of cognitively healthy adults without conditions known to affect hippocampal morphometry. The full exclusion flow is shown in @fig-prisma.

::: {#fig-prisma}

```{=latex}
\centering
\begin{tikzpicture}[
  node distance=0.8cm and 1.6cm,
  every node/.style={font=\small},
  process/.style={rectangle, draw=black!70, thick, text width=4.2cm,
                  minimum height=0.9cm, align=center, rounded corners=3pt},
  decision/.style={rectangle, draw=black!70, thick, text width=3.2cm,
                   minimum height=0.9cm, align=center, rounded corners=3pt,
                   fill=black!5},
  excluded/.style={rectangle, draw=black!70, thick, text width=3.2cm,
                   minimum height=0.9cm, align=center, rounded corners=3pt,
                   fill=black!8, text=black!70},
  sample/.style={rectangle, draw=black!70, very thick, text width=4.2cm,
                 minimum height=0.9cm, align=center, rounded corners=3pt},
  subsample/.style={rectangle, draw=black!70, thick, text width=3.6cm,
                    minimum height=0.9cm, align=center, rounded corners=3pt,
                    fill=black!3, dashed},
  arrow/.style={-{Stealth[length=5pt]}, thick, black!70}
]
  % Main flow (top to bottom)
  \node[process] (initial)
    {UK Biobank Imaging Cohort\\[2pt]\textbf{N\,=\,47,398}};

  \node[decision, below=of initial] (qc)
    {Quality control filter};

  \node[process, below=of qc] (passed)
    {Passed all processing\\[2pt]\textbf{n\,=\,38,567}};

  \node[decision, below=of passed] (gq)
    {ICD-10 exclusion:\\G00--G99, Q00--Q07};

  \node[sample, below=of gq] (sensitivity)
    {Sensitivity sample\\[2pt]\textbf{N\,=\,30,032}};

  \node[decision, below=of sensitivity] (f)
    {ICD-10 exclusion:\\F00--F99};

  \node[sample, below=of f] (primary)
    {\textbf{Primary sample}\\[2pt]\textbf{N\,=\,27,680}};

  % Exclusion boxes (right side)
  \node[excluded, right=of qc] (exc1)
    {Excluded\\[2pt]n\,=\,8,831};

  \node[excluded, right=of gq] (exc2)
    {Excluded\\[2pt]n\,=\,8,535};

  \node[excluded, right=of f] (exc3)
    {Excluded\\[2pt]n\,=\,2,352};

  % Subsamples (below primary, three across)
  \node[subsample, below=1.2cm of primary, xshift=-4.2cm] (england)
    {England only\\[2pt]\textit{N\,=\,25,172}\\[-1pt]{\scriptsize(SEM analyses)}};

  \node[subsample, below=1.2cm of primary] (matched)
    {ICV-matched\\[2pt]\textit{N\,=\,9,500}\\[-1pt]{\scriptsize(Sex differences)}};

  \node[subsample, below=1.2cm of primary, xshift=4.2cm] (longitudinal)
    {Longitudinal\\[2pt]\textit{N\,=\,2,935}\\[-1pt]{\scriptsize(Normative calibration)}};

  % Arrows
  \draw[arrow] (initial) -- (qc);
  \draw[arrow] (qc) -- (passed);
  \draw[arrow] (qc) -- (exc1);
  \draw[arrow] (passed) -- (gq);
  \draw[arrow] (gq) -- (sensitivity);
  \draw[arrow] (gq) -- (exc2);
  \draw[arrow] (sensitivity) -- (f);
  \draw[arrow] (f) -- (primary);
  \draw[arrow] (f) -- (exc3);
  \draw[arrow, dashed] (primary.south) -- ++(0,-0.35cm) -| (england.north);
  \draw[arrow, dashed] (primary.south) -- (matched.north);
  \draw[arrow, dashed] (primary.south) -- ++(0,-0.35cm) -| (longitudinal.north);
\end{tikzpicture}
```

**Participant selection flow.** Dashed boxes indicate analytic subsamples derived from the primary sample.
:::

### Quality Control

Image segmentations underwent quality control in three stages: (1) automated quality scores from deep learning-based assessments, requiring DARQ score $\geq$ 0.25 [@fonov_darq_2022] and AssemblyNet rating of "A" (highest quality grade) [@coupe_assemblynet_2020], (2) statistical outlier detection excluding participants with age- and head-size-adjusted volumes exceeding $\pm 3$ SD from the sample mean, and (3) visual inspection of flagged cases by trained raters for evident segmentation failures, motion artifacts, or anatomical anomalies.

For participants with multiple imaging sessions, QC filtering was applied first, then one session per participant was selected for cross-sectional analyses. When a participant's baseline session (ses-2) failed QC but the follow-up session (ses-3) passed, the follow-up session was retained in the cross-sectional sample. Longitudinal analyses required both sessions to pass QC.

## Brain Measures

### Image Acquisition

Images were acquired on Siemens Skyra 3T scanners at three UK sites (Cheadle, Newcastle, Reading) using a T1-weighted MPRAGE sequence at 1mm isotropic resolution. Full protocol details are provided in @alfaro-almagro_image_2018.

### Segmentation

Hippocampal and ventricular volumes were segmented using an in-house convolutional neural network trained on manually labeled data. Validation against manual segmentations showed Dice $\kappa$ = 0.94 and Pearson r = 0.99 [@fernandez-lozano_enhanced_2025]. Outputs included bilateral hippocampal volumes (left, right, total) and lateral ventricular volumes (left, right, total). Intracranial volume (ICV) was obtained from AssemblyNet [@coupe_assemblynet_2020], a large-ensemble convolutional neural network for whole-brain MRI segmentation.

### Head-Size Adjustment Methods

Four adjustment approaches were applied to hippocampal and ventricular volumes [@sanchis-segura_effects_2020; @sanfilipo_correction_2004]:

1. **Unadjusted**: raw volumes in native space (mL), which preserve natural variation but are confounded by overall head size.

2. **Proportions**: volume expressed as a proportion of ICV, which assumes isometric (1:1) scaling and may over- or under-correct depending on the true allometric relationship [@de_jong_allometric_2017; @williams_neuroanatomical_2021]:

$$V_{\text{PRP}} = \frac{V}{\text{ICV}} \times 1000$$ {#eq-prp}

3. **Stereotaxic**: volumes in standard template space after nonlinear registration, which corrects for global size and shape differences:

$$V_{\text{STX}} = V \times |J|$$ {#eq-stx}

where $J$ is the Jacobian determinant of the nonlinear transformation.

4. **Residualized**: linear regression residuals after removing ICV association, with regression coefficients estimated from the pooled sample [@sanchis-segura_effects_2020; @obrien_statistical_2011]:

$$V_{\text{RES}} = V - \hat{\beta} \times (\text{ICV} - \overline{\text{ICV}})$$ {#eq-res}

Residualization eliminates ICV-related variance and produces more replicable sex differences compared to proportional methods.

### HVR Calculation

The hippocampal-to-ventricle ratio (HVR) was computed as:

$$\text{HVR} = \frac{\text{HC}_{\text{total}}}{\text{HC}_{\text{total}} + \text{LV}_{\text{total}}}$$ {#eq-hvr}

where $\text{HC}_{\text{total}}$ and $\text{LV}_{\text{total}}$ denote bilateral sums (left + right) of hippocampal and lateral ventricular volumes, respectively. HVR is bounded between 0 and 1, with higher values indicating larger hippocampus relative to ventricles. Because HVR is a ratio of two head-size-dependent structures, no further head-size adjustment was applied.

## Cognitive Assessment

### Test Battery

Cognitive data were collected at the imaging visit using a touchscreen-based battery of 12 indicators spanning memory, processing speed, executive function, and reasoning domains (`r S("tbl-cog-battery")`):

1. **Pair matching -- average time** (mean response time; inverse-coded)
2. **Pair matching -- average errors** (mean incorrect responses; inverse-coded)
3. **Numeric memory** (maximum digits recalled)
4. **Prospective memory** (recall timing: 0 = incorrect, 1 = second attempt, 2 = first attempt)
5. **Fluid intelligence** (total score)
6. **Matrix pattern recognition** (puzzles solved)
7. **Trail making -- numeric** (completion time; inverse-coded)
8. **Trail making -- alphanumeric** (completion time; inverse-coded)
9. **Tower rearranging** (correct puzzles)
10. **Reaction time** (mean time to identify matches; inverse-coded)
11. **Symbol digit substitution -- correct** (correct responses)
12. **Symbol digit substitution -- attempts** (total attempts)

Missing cognitive data were handled using full information maximum likelihood (FIML) estimation, which uses all available data under a missing-at-random assumption.

### Factor Structure: Bifactor CFA

Cognitive indicators were modeled using a bifactor confirmatory factor analysis (CFA) [@sudlow_uk_2015] with three latent factors. The general factor (*g*) loaded on all 12 indicators. The memory-specific factor (Memory~s~, orthogonal to *g*) loaded on pair matching, numeric memory, and prospective memory. The processing speed-specific factor (Speed~s~, orthogonal to *g*) loaded on reaction time, trail making numeric, and symbol digit substitution correct. Factor scales were set using the marker variable method (first loading fixed to 1.0 for each factor). All latent factors were constrained to be orthogonal, consistent with bifactor model assumptions. Cross-loadings were fixed to zero. Residual covariances were specified for indicator pairs sharing method variance within the same test (pair matching time with errors; symbol digit correct with attempts; trail making numeric with alphanumeric).

Models were estimated using maximum likelihood with robust standard errors (MLR) in lavaan [@rosseel_lavaan_2012].

## Statistical Analyses

### Analysis 1: HVR Self-Normalizing Properties

To evaluate whether HVR reduces head-size confounding, we computed Pearson correlations between ICV and each brain measure (HC, LV, HVR) as well as between ICV and each head-size adjusted HC volume (proportions, stereotaxic, residualized). We expected a self-normalizing measure to show reduced correlation with ICV compared to unadjusted volumes.

### Analysis 2: Sex Differences in Brain Measures

Sex differences were quantified using Cohen's d with 95% confidence intervals [@cohen_statistical_2013], with positive d indicating females > males by our coding convention. Effect sizes were computed for all ROI x adjustment method combinations in the full sample (N = `r format(n_primary, big.mark=",")`).

To isolate sex effects from head-size confounding, we created a matched subsample where females and males were paired on age ($\pm 1$ year) and ICV ($\pm 25$ cc, approximately 0.3 SD). Matching used the MatchIt package (v4.7.2; @ho_matchit_2011) with optimal 1:1 matching without replacement (`method = "optimal"`, `distance = "mahalanobis"`), with calipers enforcing tolerance bounds. This yielded N = `r format(n_matched_pairs, big.mark=",")` matched pairs (total N = `r format(n_matched_pairs * 2, big.mark=",")`). We reasoned that if sex differences in HC persist after matching on ICV, they cannot be attributed solely to head-size confounding, and if HVR shows similar effects in full and matched samples, this indicates HVR is independent of head size.

### Analysis 3: Age x Sex Interactions

Linear regression models tested whether cross-sectional age-related patterns differed by sex:

$$Y_i = \beta_0 + \beta_{\text{Age}} \cdot \text{Age}_i + \beta_{\text{Sex}} \cdot \text{Sex}_i + \beta_{\text{Age} \times \text{Sex}} \cdot \text{Age}_i \times \text{Sex}_i + \varepsilon_i$$ {#eq-age-sex}

where $Y$ is the brain measure, Sex is coded as a binary indicator (Female = reference), and $\varepsilon_i \sim \mathcal{N}(0, \sigma^2)$. A significant $\beta_{\text{Age} \times \text{Sex}}$ indicates different age slopes between sexes. Sex-specific slopes were derived as $\hat{\beta}_{\text{Age}}$ for females and $\hat{\beta}_{\text{Age}} + \hat{\beta}_{\text{Age} \times \text{Sex}}$ for males, and their ratio quantified the relative rate of cross-sectional age-related differences. Separate models were fit for HC~RES~ and LV~RES~ (both residualized against ICV; @eq-res) and HVR (self-normalizing; @eq-hvr). While age trajectories may be non-linear (as captured by the GAMLSS normative models below), linear interaction models were used here to provide interpretable summary statistics for sex-differential aging rates.

### Analysis 4: Normative Modeling

Sex-specific normative centile curves were estimated using Generalized Additive Models for Location, Scale, and Shape (GAMLSS) [@rigby_generalized_2005; @bethlehem_brain_2022]. For HC and LV (unadjusted):

$$\mu = f_\mu(\text{Age}) + \beta_{\text{ICV}} \cdot \text{ICV} + \beta_{\text{Edu}} \cdot \text{Education} + u_{\text{site}}$$ {#eq-gamlss-hc}

For HVR (self-normalizing, no ICV term needed):

$$\mu = f_\mu(\text{Age}) + \beta_{\text{Edu}} \cdot \text{Education} + u_{\text{site}}$$ {#eq-gamlss-hvr}

where $f_\mu(\cdot)$ denotes a cubic spline smooth for the location parameter and $u_{\text{site}}$ denotes random intercepts for scanner site. The scale parameter was also modeled as a function of age, $\sigma = f_\sigma(\text{Age})$, allowing variance to change across the age range. Model families (Gaussian and Box-Cox Cole and Green; BCCG) were compared using AIC, with BIC as secondary criterion. The BCCG distribution was used when flexible modeling of location ($\mu$), scale ($\sigma$), and skewness ($\nu$) was warranted; shape parameters beyond $\sigma$ were held constant to maintain parsimony. Centile curves (5th, 10th, 25th, 50th, 75th, 90th, 95th) were generated for ages 45--82 by sex.

Data were split into 80% training and 20% test sets, stratified by sex and 5-year age bins. Distribution families were selected on training data and evaluated on the held-out test set by comparing observed vs. expected proportions across centiles. Final models were then fit on the complete cross-sectional sample using the selected distribution family. Model calibration was additionally assessed using the longitudinal subsample (N = 2,935; see Data Source and Participants). Because this subsample comprises the same participants as the primary cross-sectional sample with an additional follow-up session, it tests temporal stability rather than independent validation. Test-retest reliability was quantified using Pearson correlation and intraclass correlation coefficient (ICC, two-way random, absolute agreement).

Two sets of GAMLSS models were generated: (1) site-controlled models including random site intercepts ($u_{\text{site}}$) for optimal within-UKB accuracy, and (2) transfer models without site effects for application to external cohorts where site is unknown or not comparable. Transfer models used identical distribution families but omitted the random site term, following recommendations for cross-study normative modeling.

### Analysis 5: Brain-Cognition SEM and Measurement Invariance

Joint measurement-structural equation models tested associations between brain measures and cognitive factors. These analyses were restricted to the England subsample (N = `r format(n_england, big.mark=",")`; see Data Source and Participants) because the Index of Multiple Deprivation (IMDP), used as a socioeconomic covariate, is not comparable across UK countries. All models were estimated in lavaan [@rosseel_lavaan_2012].

#### Measurement Invariance

Comparing brain-cognition associations across sexes requires that the cognitive measurement model is equivalent across groups. Measurement invariance of the bifactor structure (see Factor Structure: Bifactor CFA) was therefore tested following standard multi-group CFA procedures [@putnick_measurement_2016]. Four nested models were tested: (1) configural invariance (same factor structure), (2) metric invariance (equal loadings), (3) scalar invariance (equal intercepts), and (4) strict invariance (equal residuals). Invariance was evaluated using $\Delta$CFI $\leq$ 0.01 [@cheung_evaluating_2002] and $\Delta$RMSEA $\leq$ 0.015 [@chen_sensitivity_2007]. At minimum, metric invariance is required to compare structural paths across groups.

#### SEM Specifications

Given adequate measurement invariance, three SEM models were estimated, each combining the bifactor cognitive measurement model with structural paths from a brain measure to all three cognitive factors. Brain measures entered as observed (manifest) variables: HC (bilateral hippocampal volume, sum of left and right), HVR (@eq-hvr), and HC~RES~ (HC residualized against ICV within sex, yielding r $\approx$ 0 with ICV by construction). The structural component of each model took the following form:

**Model 1 (HC):**

$$\begin{aligned}
\text{HC} &= \beta_{\text{Age}} \cdot \text{Age} + \beta_{\text{ICV}} \cdot \text{ICV} + \beta_{\text{Edu}} \cdot \text{Edu} + \beta_{\text{IMDP}} \cdot \text{IMDP} + \zeta_{\text{HC}} \\
\eta_k &= \beta_{k,\text{HC}} \cdot \text{HC} + \beta_{k,\text{Age}} \cdot \text{Age} + \beta_{k,\text{ICV}} \cdot \text{ICV} + \beta_{k,\text{Edu}} \cdot \text{Edu} + \beta_{k,\text{IMDP}} \cdot \text{IMDP} + \zeta_k
\end{aligned}$$ {#eq-sem-hc}

**Model 2 (HVR):**

$$\begin{aligned}
\text{HVR} &= \beta_{\text{Age}} \cdot \text{Age} + \beta_{\text{Age}^2} \cdot \text{Age}^2 + \beta_{\text{ICV}} \cdot \text{ICV} \\
  &\quad + \beta_{\text{Edu}} \cdot \text{Edu} + \beta_{\text{IMDP}} \cdot \text{IMDP} + \zeta_{\text{HVR}} \\
\eta_k &= \beta_{k,\text{HVR}} \cdot \text{HVR} + \beta_{k,\text{Age}} \cdot \text{Age} + \beta_{k,\text{Age}^2} \cdot \text{Age}^2 \\
  &\quad + \beta_{k,\text{ICV}} \cdot \text{ICV} + \beta_{k,\text{Edu}} \cdot \text{Edu} + \beta_{k,\text{IMDP}} \cdot \text{IMDP} + \zeta_k
\end{aligned}$$ {#eq-sem-hvr}

**Model 3 (HC~RES~):**

$$\begin{aligned}
\text{HC}_{\text{RES}} &= \beta_{\text{Age}} \cdot \text{Age} + \beta_{\text{Age}^2} \cdot \text{Age}^2 + \beta_{\text{ICV}} \cdot \text{ICV} \\
  &\quad + \beta_{\text{Edu}} \cdot \text{Edu} + \beta_{\text{IMDP}} \cdot \text{IMDP} + \zeta_{\text{HC}_{\text{RES}}} \\
\eta_k &= \beta_{k,\text{HC}_{\text{RES}}} \cdot \text{HC}_{\text{RES}} + \beta_{k,\text{Age}} \cdot \text{Age} + \beta_{k,\text{Age}^2} \cdot \text{Age}^2 \\
  &\quad + \beta_{k,\text{ICV}} \cdot \text{ICV} + \beta_{k,\text{Edu}} \cdot \text{Edu} + \beta_{k,\text{IMDP}} \cdot \text{IMDP} + \zeta_k
\end{aligned}$$ {#eq-sem-hcres}

where $\eta_k \in \{g, \text{Memory}_s, \text{Speed}_s\}$ denotes latent cognitive factors and $\zeta$ denotes residual terms. A quadratic age term was included in the HVR and HC~RES~ models but not in the HC model, where the combination of raw hippocampal volume, ICV, and quadratic age produced multicollinearity. Age was centered within sex groups before squaring. All structural variables were z-standardized within sex groups to address scale differences. Covariates are summarized in `r S("tbl-sem-covariates-fields")`.

The HC~RES~ model serves as a benchmark representing the methodological standard for removing head-size confounding. Comparing HVR to HC~RES~ tests whether HVR provides similar cognitive associations while being self-normalizing and capturing additional variance through its incorporation of ventricular dynamics. The proportions and stereotaxic adjustment methods were excluded due to concerns about biologically implausible associations and residual ICV correlation, respectively [@wang_comparison_2024].

#### Estimation and Inference

Two estimation approaches were used. Pooled models combined all participants, included sex as a covariate, and were estimated using robust maximum likelihood (MLR) with cluster-robust standard errors to account for scanner site. Multi-group models estimated separate structural parameters for females and males, allowing sex differences in brain-cognition paths to be directly tested; these were estimated using maximum likelihood with bootstrap confidence intervals (2,000 resamples) [@cumming_inference_2005]. Missing cognitive data were handled using full information maximum likelihood (FIML) in both approaches. Sex differences in path coefficients were evaluated by comparing bootstrap confidence intervals across groups.

### Analysis 6: Sensitivity Analyses

Three sensitivity analyses assessed the robustness of sex difference findings to analytical choices.

#### Site effects

Scanner site (UK Biobank assessment centre) can introduce systematic variance in volumetric measurements due to differences in scanner hardware, software versions, or acquisition protocols. Site effects were quantified using the intraclass correlation coefficient:

$$\text{ICC} = \frac{\sigma^2_{\text{site}}}{\sigma^2_{\text{site}} + \sigma^2_{\text{residual}}}$$ {#eq-icc}

Site-adjusted sex difference effect sizes were computed using linear mixed-effects models with a fixed effect for sex and a random intercept for scanner site:

$$Y_i = \beta_0 + \beta_{\text{Sex}} \cdot \text{Sex}_i + u_{\text{site}(i)} + \varepsilon_i$$ {#eq-site-adj}

where $u_{\text{site}} \sim \mathcal{N}(0, \sigma^2_{\text{site}})$ and $\varepsilon_i \sim \mathcal{N}(0, \sigma^2_{\text{residual}})$. Cohen's d was approximated as $\hat{\beta}_{\text{Sex}} / \hat{\sigma}_{\text{residual}}$, with confidence intervals derived from the Wald approximation.

#### Hemisphere comparison

Left and right hippocampus and lateral ventricle volumes (residualized against ICV) were analyzed separately to assess whether sex differences are lateralized. Cohen's d was computed for each hemisphere independently, with effect size comparisons based on overlapping 95% confidence intervals. Non-overlapping intervals indicate a statistically significant difference in effect magnitude between hemispheres.

#### Inclusion of psychiatric diagnoses

The primary analysis excluded participants with ICD-10 diagnoses in the F (mental and behavioral disorders), G (neurological), or Q0 (congenital brain malformations) categories. To assess robustness, sex difference analyses were repeated in an expanded sensitivity sample that included participants with F-code diagnoses (F00-F99) while still excluding G and Q0 codes. The difference in Cohen's d between primary and sensitivity samples ($\Delta d$ = $d_{\text{primary}}$ - $d_{\text{sensitivity}}$) quantified the impact of psychiatric exclusion criteria. Small differences indicate that findings generalize to a less restrictively selected sample and are not driven by exclusion of psychiatric comorbidity.

---

# Results

## Sample Characteristics

The primary sample comprised `r format(n_primary, big.mark=",")` participants (`r format(n_female, big.mark=",")` female, `r format(n_male, big.mark=",")` male; `r pct_female`% female) aged `r age_min`--`r age_max` years (@tbl-demographics). Females were slightly younger on average (M = `r age_mean_f`, SD = `r age_sd_f`) than males (M = `r age_mean_m`, SD = `r age_sd_m`). Males had substantially larger ICV than females, consistent with the well-documented sex difference in head size. The England subsample (N = `r format(n_england, big.mark=",")`), ICV-matched subsample (N = `r format(n_matched_pairs * 2, big.mark=",")`; `r format(n_matched_pairs, big.mark=",")` pairs), and longitudinal subsample (N = 2,935) showed comparable age and ICV distributions to the primary sample within their respective selection constraints.

```{r}
#| label: tbl-demographics
#| tbl-cap: "**Participant characteristics by sex and sample.** Values are M (SD). d = Cohen's d (Female − Male). England = English assessment centres (SEM analyses). Matched = age $\\pm 1$ yr, ICV $\\pm 25$ cc. Longitudinal = participants with repeat imaging."

lng_demog <- if ("LNG" %in% names(demog_data)) demog_data$LNG else NULL
format_demographics_gt(all_demog, eng_demog, mtch_demog, lng_demog)
```

## HVR Self-Normalizing Properties (Analysis 1)

Unadjusted HC showed strong positive correlation with ICV (r = `r round(hvr_icv[VARIABLE == "HC", CORRELATION], 3)`), while proportions-adjusted HC showed moderate negative correlation (r = `r round(hvr_icv[VARIABLE == "HC_PRP", CORRELATION], 3)`), indicating over-correction (@tbl-hvr-validation). Residualization eliminated ICV association by construction (r $\approx$ 0). HVR's correlation with ICV (r = `r round(hvr_icv[VARIABLE == "HVR", CORRELATION], 3)`) was smaller in magnitude than raw HC (r = `r round(hvr_icv[VARIABLE == "HC", CORRELATION], 3)`) but remained negative, unlike residualized HC (r $\approx$ 0). This represents a `r round((1 - abs(hvr_icv[VARIABLE == "HVR", CORRELATION]) / abs(hvr_icv[VARIABLE == "HC", CORRELATION])) * 100, 0)`% reduction in absolute ICV association. The negative direction mirrors the LV-ICV relationship (r = `r round(hvr_icv[VARIABLE == "LV", CORRELATION], 3)`), indicating that the HVR-ICV association reflects the ventricular component's scaling with head size.

```{r}
#| label: tbl-hvr-validation
#| tbl-cap: "**Brain measure correlations with ICV.** An ideal self-normalizing measure shows r(ICV) near zero."

format_hvr_validation_gt(hvr_icv)
```

## Sex Differences in Brain Volumes (Analysis 2)

Sex differences varied across brain measures and adjustment methods (N = `r format(n_primary, big.mark=",")`; @fig-effect-sizes).

### Hippocampal Volume

HC sex differences reversed direction depending on the adjustment method. Unadjusted volumes showed d = `r round(hc_non$ESTIMATE, 2)` [95% CI: `r round(hc_non$CI_LOWER, 2)`, `r round(hc_non$CI_UPPER, 2)`], indicating males had larger absolute volumes. Proportions adjustment reversed this to d = `r round(hc_prp$ESTIMATE, 2)` [`r round(hc_prp$CI_LOWER, 2)`, `r round(hc_prp$CI_UPPER, 2)`] with females larger. Stereotaxic adjustment yielded d = `r round(hc_stx$ESTIMATE, 2)` [`r round(hc_stx$CI_LOWER, 2)`, `r round(hc_stx$CI_UPPER, 2)`], and residualized volumes showed a negligible effect (d = `r round(hc_res$ESTIMATE, 2)` [`r round(hc_res$CI_LOWER, 2)`, `r round(hc_res$CI_UPPER, 2)`]). The effect size range across methods spanned `r round(abs(min(bilateral[ROI == "HC"]$ESTIMATE) - max(bilateral[ROI == "HC"]$ESTIMATE)), 2)` standard deviations.

### Lateral Ventricular Volume

Unlike HC, lateral ventricular sex differences were consistently negative (males > females) across all adjustment methods, though effect magnitudes were attenuated by head-size correction. Unadjusted LV showed d = `r round(lv_non$ESTIMATE, 2)` [`r round(lv_non$CI_LOWER, 2)`, `r round(lv_non$CI_UPPER, 2)`], similar in magnitude to unadjusted HC. After adjustment, all methods retained the male advantage: proportions d = `r round(lv_prp$ESTIMATE, 2)` [`r round(lv_prp$CI_LOWER, 2)`, `r round(lv_prp$CI_UPPER, 2)`], stereotaxic d = `r round(lv_stx$ESTIMATE, 2)` [`r round(lv_stx$CI_LOWER, 2)`, `r round(lv_stx$CI_UPPER, 2)`], and residualized d = `r round(lv_res$ESTIMATE, 2)` [`r round(lv_res$CI_LOWER, 2)`, `r round(lv_res$CI_UPPER, 2)`]. The directional consistency across methods contrasts with the reversals observed for HC.

### HVR

HVR showed a consistent female advantage: d = `r round(hvr_non$ESTIMATE, 2)` [95% CI: `r round(hvr_non$CI_LOWER, 2)`, `r round(hvr_non$CI_UPPER, 2)`]. The combination of males having relatively smaller hippocampi (after head-size adjustment) and larger ventricles produces this female advantage in HVR. Distribution overlap between sexes across adjustment methods is shown in `r S("fig-distributions")`.

```{r}
#| label: fig-effect-sizes
#| fig-cap: "**Sex differences by adjustment method.** Circles = full sample; triangles = ICV-matched sample. Faded points indicate non-significant effects (95% confidence interval crosses zero). Right panel shows $\\Delta d$ = $d_{\\text{full}}$ $-$ $d_{\\text{matched}}$, representing the change in effect size after removing head-size confounding; values near zero indicate head-size-independent effects."
#| fig-width: 6.5
#| fig-height: 5.6
#| dev: cairo_pdf
fig_effect_sizes_by_adjustment(sex_diff, include_matched = TRUE)
```

### Matched-Pair Comparison

Matching on age and ICV (N = `r format(n_matched_pairs * 2, big.mark=",")`; `r format(n_matched_pairs, big.mark=",")` pairs) revealed which sex differences persist after removing head-size confounding (@tbl-matched). HC effects collapsed to near-zero (unadjusted d = `r round(matched_hc_non$ESTIMATE_MTCH, 2)`), confirming that HC sex differences are driven by head-size variation. LV effects were attenuated but remained significant (unadjusted d = `r round(matched_lv_non$ESTIMATE_MTCH, 2)`), indicating that males have larger ventricles even at equivalent head size. HVR was reduced but persisted (d = `r round(matched_hvr$ESTIMATE_MTCH, 2)` [`r round(matched_hvr$CI_LOWER_MTCH, 2)`, `r round(matched_hvr$CI_UPPER_MTCH, 2)`]), consistent with the female HVR advantage being independent of head-size confounding.

```{r}
#| label: tbl-matched
#| tbl-cap: "**Full vs. matched sample effect sizes.** Matching on age ($\\pm 1$ yr) and ICV ($\\pm 25$ cc) eliminates head-size confounding."
if ("COMPARISON" %in% names(sex_diff))
  format_matched_comparison_gt(sex_diff$COMPARISON, n_matched_pairs)
```

## Age x Sex Interactions (Analysis 3)

Cross-sectional age patterns (N = `r format(n_primary, big.mark=",")`) differed by sex (@fig-age-trajectory). For HVR, the age × sex interaction was significant ($\beta$ = `r round(hvr_int$INTERACTION_EST, 4)`, SE = `r round(hvr_int$INTERACTION_SE, 5)`, t = `r round(hvr_int$INTERACTION_T, 2)`, p = `r format(hvr_int$INTERACTION_P, scientific=TRUE, digits=2)`), with males showing `r round(abs(hvr_int$SLOPE_RATIO), 2)`× steeper cross-sectional age-related differences. The female HVR advantage *widened* with age—effect sizes increased from d = `r round(hvr_d_young$ESTIMATE, 2)` at ages 45-55 to d = `r round(hvr_d_old$ESTIMATE, 2)` at ages 75+.

Age × sex interactions were significant for all three measures. HC showed males with `r round(abs(hc_int$SLOPE_RATIO), 2)`× steeper cross-sectional differences (p = `r format(hc_int$INTERACTION_P, scientific=TRUE, digits=2)`). LV showed the largest sex disparity, with males exhibiting `r round(abs(lv_int$SLOPE_RATIO), 2)`× steeper ventricular expansion (p = `r format(lv_int$INTERACTION_P, scientific=TRUE, digits=2)`). The greater sex disparity in ventricular trajectories compared to hippocampal trajectories explains the intermediate HVR ratio (`r round(abs(hvr_int$SLOPE_RATIO), 2)`×), as HVR captures both processes.

```{r}
#| label: fig-age-trajectory
#| fig-cap: "**Age trajectories of sex differences.** Solid line = hippocampal-to-ventricle ratio; dashed lines = hippocampal/ventricular volume. White-filled points = non-significant (95% confidence interval crosses zero)."
#| fig-width: 6.5
#| fig-height: 4.5
fig_age_trajectory(sex_diff)
```

## Normative Modeling (Analysis 4)

@fig-normative-centiles presents sex-specific normative centiles for HVR, hippocampal volume, and lateral ventricular volume. Across the age range, females showed higher HVR values at all percentile levels; the median female HVR at age 55 corresponded approximately to the 75th percentile for males of the same age. HC and LV centile curves are generated from GAMLSS models that include ICV as a covariate, providing head-size adjusted norms without pre-residualizing the data. Males show greater ventricular variability at older ages, potentially reflecting greater heterogeneity in aging trajectories.

```{r}
#| label: fig-normative-centiles
#| fig-cap: "**Sex-specific normative centile curves.** Hippocampal-to-ventricle ratio (top), hippocampal volume (middle), and lateral ventricular volume (bottom). Left and right columns show female and male centiles separately; the centre column overlays both sexes for direct comparison. Hippocampal and ventricular volume models include intracranial volume as a covariate."
#| fig-width: 6.5
#| fig-height: 8
fig_normative_centiles_panel(norm_tables)
```

Calibration plots (`r S("fig-gamlss-calibration")`) demonstrate that normative models accurately capture the distribution of observed values, with observed proportions closely tracking expected proportions across centiles.

### Temporal Stability (Longitudinal Validation)

All brain measures demonstrated high test-retest reliability across the longitudinal follow-up (`r S("tbl-temporal-stability")`), with Pearson correlations > 0.95 and ICCs > 0.93 for all measures. HVR showed slightly lower reliability (r = `r round(hvr_r, 3)`, ICC = `r round(hvr_icc, 3)`) compared to HC (r = `r round(hc_r, 3)`, ICC = `r round(hc_icc, 3)`) and LV (r = `r round(lv_r, 3)`, ICC = `r round(lv_icc, 3)`), consistent with the ratio measure amplifying small measurement errors in either component. These reliabilities support the use of these normative models for individual-level assessment.

### Clinical Normative Centile Tables

Sex-specific normative centile tables for HVR, hippocampal volume, and lateral ventricular volume are provided in Supporting Information (`r S_range("tbl-hvr-centiles-female", "tbl-lv-centiles-male")`) to enable clinical interpretation of individual measurements at 5-year age intervals. These tables allow clinicians to determine an individual's percentile rank given their age, sex, and brain measure.

GAMLSS model coefficients for the transfer models (excluding site random effects) are provided in `r S("tbl-gamlss-hvr-coef")` and `r S("tbl-gamlss-hc-coef")`. These transfer models can be applied to external datasets for computing individual centile scores.

## Brain-Cognition Associations (Analysis 5)

SEM analyses were restricted to the England subsample (N = `r format(n_england, big.mark=",")`) for comparability of the Index of Multiple Deprivation (IMDP) covariate.

### Measurement Invariance

The cognitive bifactor model showed acceptable fit at all invariance levels, with $\Delta$CFI $\leq$ 0.01 and $\Delta$RMSEA $\leq$ 0.015 at each step (`r S("tbl-measurement-invariance")`). Strict measurement invariance was supported, validating cross-sex comparisons of factor scores and structural paths.

### Pooled Model Results

General cognition (*g*) was positively associated with all three brain measures (@tbl-sem-paths): unadjusted HC ($\beta$ = `r safe_round(hc_g$est.std, 3)` [95% CI: `r safe_round(hc_g$ci.lower, 3)`, `r safe_round(hc_g$ci.upper, 3)`]), HVR ($\beta$ = `r safe_round(hvr_g$est.std, 3)` [`r safe_round(hvr_g$ci.lower, 3)`, `r safe_round(hvr_g$ci.upper, 3)`]), and HC~RES~ ($\beta$ = `r safe_round(hc_res_g$est.std, 3)` [`r safe_round(hc_res_g$ci.lower, 3)`, `r safe_round(hc_res_g$ci.upper, 3)`]), all p < .001. Confidence intervals overlapped substantially, indicating no significant differences in predictive validity between measures. HVR showed numerically the strongest association with *g*, despite requiring no head-size adjustment.

For processing speed (Speed~s~), HC showed the strongest effect ($\beta$ = `r safe_round(hc_speed$est.std, 3)`, p < .001), followed by HC~RES~ ($\beta$ = `r safe_round(hc_res_speed$est.std, 3)`, p = `r safe_round(hc_res_speed$pvalue, 3)`), with HVR trending but not significant ($\beta$ = `r safe_round(hvr_speed$est.std, 3)`, p = `r safe_round(hvr_speed$pvalue, 3)`). The memory-specific factor (Memory~s~) could not be reliably estimated in either sex and is therefore not reported here (see `r S("tbl-sem-loadings")` for factor loadings).

```{r}
#| label: tbl-sem-paths
#| tbl-cap: "**Pooled brain-cognition path coefficients.** Beta = standardized path coefficient; CI from 2000 bootstrap resamples. Memory-specific factor paths not shown due to factor collapse."

format_pooled_sem_paths_gt(sem_params, hvr_g, hc_res_g, hc_g)
```

### Sex-Stratified Results

@fig-sem-forest displays sex-stratified path coefficients from multi-group models (`r S("tbl-sem-paths-by-sex")`). The multigroup analyses revealed sex differences in brain-cognition associations.

For general cognition (*g*), HC~RES~ → *g* was not significant in males (p = `r safe_round(hc_res_g_m$pvalue, 3)`) while remaining significant in females. In contrast, HVR showed the opposite pattern: males exhibited stronger HVR-cognition associations ($\beta$ = `r safe_round(hvr_g_m$est.std, 3)`) than females ($\beta$ = `r safe_round(hvr_g_f$est.std, 3)`), though both were significant. For processing speed (Speed~s~), males showed significant HC and HC~RES~ associations ($\beta$ = `r safe_round(hc_speed_m$est.std, 3)` and `r safe_round(hc_res_speed_m$est.std, 3)`, respectively) while females showed none, suggesting sex-differential neural substrates for processing speed performance. Neither sex showed significant HVR → Speed~s~ associations.

```{r}
#| label: fig-sem-forest
#| fig-cap: "**Brain-cognition paths by sex.** Points represent standardized coefficients; lines show 95% bootstrap confidence intervals. White-filled points indicate non-significant effects."
#| fig-width: 6.5
#| fig-height: 4.4
fig_sem_forest(sem_params)
```

All SEM models demonstrated good fit according to conventional criteria (`r S("tbl-sem-fit")`; @hu_cutoff_1999). Full model parameters including factor loadings and covariate effects are reported in `r S("tbl-sem-loadings")` and `r S("tbl-sem-covariate-effects")`.

#### Non-Linear Age Effects

Quadratic age terms were included in HVR and HC~RES~ models to capture non-linear cross-sectional age patterns. Raw hippocampal volume (HC) models were fit without the age² term because the combination of HC, ICV, and age² produced multicollinearity (information matrix not invertible); fit indices for the HC model (`r S("tbl-sem-fit")`) therefore reflect this linear-age specification. Both brain measures and cognitive factors showed significant negative quadratic age effects (all p < .001; `r S("tbl-sem-covariate-effects")`), indicating steeper cross-sectional differences at older ages. For brain measures, HVR showed $\beta$ = `r safe_round(hvr_age_sq$est.std, 3)` and HC~RES~ showed $\beta$ = `r safe_round(hc_res_age_sq$est.std, 3)`. For cognitive factors, *g* showed $\beta$ = `r safe_round(g_age_sq_hvr$est.std, 3)` in the HVR model and $\beta$ = `r safe_round(g_age_sq_hc$est.std, 3)` in the HC~RES~ model, while Speed~s~ showed $\beta$ = `r safe_round(speed_age_sq_hvr$est.std, 3)` and $\beta$ = `r safe_round(speed_age_sq_hc$est.std, 3)` respectively. The stronger quadratic effect for HVR compared to HC~RES~ suggests that HVR captures both hippocampal atrophy and ventricular expansion, both of which accelerate with age.

## Sensitivity Analyses (Analysis 6)

### Site-Adjusted Effect Sizes

Site ICCs ranged from 0.004 to 0.012 (`r S("tbl-site-adjusted")`), indicating that scanner site explained less than 1.5% of total variance in all brain measures. Effect sizes remained consistent after site adjustment.

### Hemisphere Comparison

Sex difference effect sizes were largely symmetrical across hemispheres for the individual volumes (`r S("tbl-hemisphere")`, `r S("fig-hemisphere")`). Residualized hippocampal volume showed significant effects bilaterally (left d = −0.14, right d = −0.16) with overlapping confidence intervals, indicating no lateralized sex difference. Lateral ventricles similarly showed a male advantage in both hemispheres with nearly identical effect sizes (left d = 0.48, right d = 0.48). However, HVR showed a lateralized pattern: the right hemisphere exhibited a significant female advantage (d = 0.05 [0.03, 0.07]), while the left hemisphere effect was non-significant (d = −0.01 [−0.03, 0.02]).

### Inclusion of Psychiatric Diagnoses (F-Codes)

Effect sizes remained stable when including participants with psychiatric diagnoses (`r S("tbl-sensitivity")`, `r S("fig-sensitivity")`), demonstrating robustness of findings to exclusion criteria.

---

# Discussion

## Summary of Findings

In `r format(n_primary, big.mark=",")` UK Biobank participants, we evaluated HVR as a self-normalizing biomarker that bypasses conventional head-size correction. HVR showed weak negative correlation with ICV (r = `r round(r_hvr_icv, 3)`), representing a `r icv_reduction_pct`% reduction in ICV dependence compared to unadjusted HC; the residual negative correlation may reflect ventricular scaling rather than incomplete normalization. Hippocampal sex differences ranged from d = `r round(hc_non$ESTIMATE, 2)` (males larger) to d = `r round(hc_prp$ESTIMATE, 2)` (females larger) depending solely on the adjustment method, whereas HVR consistently showed females higher than males (d = `r round(hvr_non$ESTIMATE, 2)`), an effect replicated in ICV-matched samples. HVR predicted general cognitive ability comparably to hippocampal volume (overlapping confidence intervals). Males showed steeper age-related HVR differences (`r round(abs(hvr_int$SLOPE_RATIO), 2)`× female slope), with the female advantage widening at older ages (d = `r round(hvr_d_young$ESTIMATE, 2)` at 45-55y to d = `r round(hvr_d_old$ESTIMATE, 2)` at 75+y).

## HVR Self-Normalizing Properties (Analysis 1)

HVR achieves head-size independence through its mathematical construction: both numerator (HC) and denominator (HC + LV) scale with head size, so their ratio largely normalizes out this confound [@schoemaker_hippocampal-ventricle_2019]. In two healthy aging cohorts (N = 50 and N = 88), @schoemaker_hippocampal-ventricle_2019 demonstrated that HVR yielded stronger correlations with both age and memory performance than absolute hippocampal volume, supporting its utility as an integrity index. In the present sample, HVR achieved a `r icv_reduction_pct`% reduction in ICV correlation compared to unadjusted HC (r = `r round(r_hvr_icv, 3)` vs. r = `r round(r_hc_icv, 3)`). This self-normalizing approach complements other linear ratio measures used in clinical neuroimaging, such as the Evans' index for ventricular assessment [@curra_ventricular_2019], and avoids the well-documented pitfalls of proportional ICV correction, which produces biologically implausible results when brain structures scale non-isometrically with head size [@wang_comparison_2024; @pintzka_marked_2015; @sanchis-segura_effects_2020].

The residual negative correlation between HVR and ICV (r = `r round(r_hvr_icv, 3)`) likely reflects ventricular scaling rather than incomplete normalization. Lateral ventricular volume shows positive correlation with ICV (r = `r round(r_lv_icv, 3)`), and the relative size of the ventricles increases with increasing ICV [@pintzka_marked_2015]. This disproportionate scaling is confirmed by allometric analyses showing that hippocampal volume scales hypoallometrically with ICV ($\alpha$ = 0.60), while the lateral ventricles scale near-isometrically ($\alpha$ $\approx$ 1.0; @de_jong_allometric_2017). Since HVR = HC/(HC+LV), and larger heads have disproportionately larger ventricles relative to hippocampal volume, larger ICV naturally produces lower HVR. The negative HVR-ICV correlation thus mirrors the positive LV-ICV relationship, capturing the non-isometric scaling of ventricular volume with head size. Unlike residualization, which removes *all* ICV association by construction (r $\approx$ 0), HVR preserves morphological variation related to how ventricular space scales with overall brain size. This may be advantageous for applications where the distinction between "large head with typical ventricles" and "large head with enlarged ventricles" carries clinical significance.

## Sex Differences in Brain Measures (Analysis 2)

### Hippocampal Sex Differences as Methodological Artifacts

The reversal of HC sex differences across adjustment methods indicates that these differences reflect head-size confounding rather than underlying biology. Effect sizes ranged from d = `r round(hc_non$ESTIMATE, 2)` (males larger, unadjusted) to d = `r round(hc_prp$ESTIMATE, 2)` (females larger, proportions-adjusted)—a swing of `r round(abs(hc_d_range), 2)` standard deviations depending solely on analytical choices. This reversal is well-replicated: @pintzka_marked_2015 found unadjusted hippocampal volumes significantly larger in males that reversed to females larger under proportional correction; @voevodskaya_effects_2014 observed the identical pattern in the PIVUS cohort (N = 406); and @fjell_minute_2009, in a multisample study of 1,143 healthy participants, reported that unadjusted hippocampal volumes favoring males (z = −0.72) reversed to favor females when corrected for total brain volume (z = 0.13). In all cases, residualization eliminated the sex difference entirely.

This over-correction reflects allometric scaling: the medial temporal lobe (including hippocampus) scales at $\alpha$ = 0.60 relative to ICV [95% CI: 0.56-0.64], not isometrically [@de_jong_allometric_2017], and 83% of brain measures in the UK Biobank show similar non-isometric scaling [@williams_neuroanatomical_2021]. Because the regression of hippocampal volume on ICV does not pass through the origin, dividing by ICV systematically overestimates relative volume in smaller heads [@pintzka_marked_2015; @voevodskaya_effects_2014]. The near-zero effect with residualization (d = `r round(hc_res$ESTIMATE, 3)`, 95% CI crossing zero) suggests that after appropriate statistical control, hippocampal volume does not meaningfully differ between sexes.

These results converge with other large-scale studies. @ritchie_sex_2018 reported that unadjusted hippocampal sex differences in 5,216 UK Biobank participants (d $\approx$ −0.55) dropped to near-zero ($|$d$|$ < 0.03) after regression-based adjustment for total brain volume. ICV-matched UK Biobank analyses similarly show sex effects approaching zero [@brzezinski-rittner_disentangling_2025], brain-based sex classification drops from >80% to ~60% after head-size adjustment [@sanchis-segura_effects_2020], and individual brains are better characterized as mosaics of features than categorically sexed [@joel_analysis_2018]. While total hippocampal volume sex differences appear methodologically determined, sex-specific developmental trajectories in hippocampal subregions [@mu_effect_2020] and structural connectivity patterns [@ingalhalikar_sex_2014] suggest that meaningful differences may exist at finer spatial scales.

### Lateral Ventricle Sex Differences

Unlike hippocampal volume, lateral ventricles showed males > females across all adjustment methods (@fig-effect-sizes). This directional consistency is well-established: unadjusted ventricular volumes are larger in males across multiple cohorts [@pintzka_marked_2015; @voevodskaya_effects_2014; @fjell_minute_2009], and—critically—the male advantage persists or even increases after head-size correction. @fjell_minute_2009 reported that the ventricular sex difference (z = −0.32 unadjusted) actually increased after total brain volume correction (z = −0.56), and males show 28–128% greater risk of abnormal ventricular enlargement indices across age [@curra_ventricular_2019]. Under proportional adjustment, ventricles remain larger in males because—unlike the hippocampus—the LV-ICV regression has a negative y-intercept, meaning proportional correction under-corrects rather than over-corrects ventricular volumes [@pintzka_marked_2015; @voevodskaya_effects_2014]. This mathematical property explains why proportional adjustment reverses the hippocampal sex difference (positive y-intercept → over-correction) while preserving the ventricular sex difference (negative y-intercept → under-correction), with @pintzka_marked_2015 demonstrating that the sign of the y-intercept predicted the direction of proportional sex differences with 100% accuracy.

Smaller studies have reported that residualization eliminates ventricular sex differences [@voevodskaya_effects_2014; @fjell_minute_2009], though @pintzka_marked_2015 found that inferior lateral ventricle differences persisted even after residualization. In our substantially larger sample (N = `r format(n_primary, big.mark=",")`), residualized LV sex differences remained significant (d = `r round(lv_res$ESTIMATE, 2)`), suggesting that the null findings in prior studies may reflect insufficient power. The consistent male ventricular advantage across adjustment methods contrasts sharply with the reversals observed for HC, indicating that this sex difference is not a head-size artifact. Since HVR = HC/(HC+LV), the female HVR advantage arises from two converging factors: relatively larger head-size-adjusted hippocampi and relatively smaller ventricles. This dual sensitivity makes HVR particularly informative for sex differences in cross-sectional brain aging patterns.

### Matched-Pair Comparison and HVR

The matched-pair analysis (@tbl-matched) provides the strongest evidence for HVR's validity as a head-size-independent measure. Matching on age and ICV has been advocated as a "gold standard benchmark" for disentangling sex from brain size [@brzezinski-rittner_disentangling_2025], as it eliminates head-size variation without the mathematical assumptions of statistical correction [@voevodskaya_effects_2014; @nordenskjold_intracranial_2015]. When females and males were matched on age ($\pm 1$ year) and ICV ($\pm 25$ cc), the HC sex difference collapsed to d = `r round(matched_hc_non$ESTIMATE_MTCH, 3)` (effectively zero), while HVR retained a significant female advantage (d = `r round(matched_hvr$ESTIMATE_MTCH, 3)`). This hippocampal convergence replicates prior matched-sample findings: @pintzka_marked_2015 found non-significant HC differences in 152 ICV-matched pairs, and @voevodskaya_effects_2014 reported that no hippocampal differences remained in their ICV-matched subsample. Conversely, males retained significantly larger ventricles (d = `r round(matched_lv_non$ESTIMATE_MTCH, 2)`) even after ICV matching—consistent with both @pintzka_marked_2015 and @voevodskaya_effects_2014, who found that ventricular sex differences (third ventricle in both studies) survived strict ICV matching when hippocampal differences did not. The convergence of residualized HC and HVR effect sizes in matched samples provides converging evidence that both approaches successfully remove head-size confounding, though through different mechanisms: residualization statistically removes ICV variance, while HVR leverages the shared scaling of hippocampal and ventricular volumes.

The consistent female HVR advantage across full (d = `r round(hvr_non$ESTIMATE, 2)`) and matched (d = `r round(matched_hvr$ESTIMATE_MTCH, 2)`) samples supports a sex difference in the hippocampus-to-ventricle balance. The `r hvr_reduction_pct`% reduction in effect size from full to matched samples indicates that most of the HVR sex difference is attributable to head-size differences, but a significant residual effect persists. This residual could reflect greater relative hippocampal volume in females after accounting for ventricular size, smaller relative ventricular volume in females (as confirmed by the matched-sample analysis showing LV d = `r round(matched_lv_non$ESTIMATE_MTCH, 2)` favoring males even after ICV matching), or sex differences in the relationship between hippocampal size and ventricular size across ages.

## Age × Sex Interactions (Analysis 3)

The steeper male HVR differences with age suggests sex differences in age-related patterns of hippocampal and ventricular volumes. The age × sex interaction analyses revealed a consistent pattern across all measures: males showed `r round(hvr_int$SLOPE_RATIO, 2)`× steeper HVR cross-sectional differences (p = `r format(hvr_int$INTERACTION_P, scientific=TRUE, digits=2)`), `r round(hc_int$SLOPE_RATIO, 2)`× steeper HC differences (p = `r format(hc_int$INTERACTION_P, scientific=TRUE, digits=2)`), and `r round(lv_int$SLOPE_RATIO, 2)`× steeper LV differences (p = `r format(lv_int$INTERACTION_P, scientific=TRUE, digits=2)`). Steeper male subcortical volume decline is consistent with @ritchie_sex_2018, who reported significant age × sex interactions for hippocampal volume in 5,216 UK Biobank participants. For ventricular volumes, @fjell_minute_2009 found that global CSF expansion showed a steeper age correlation in males (r = 0.50) than females (r = 0.36) after ICV correction.

The greater sex disparity in ventricular age-related differences (`r round(lv_int$SLOPE_RATIO, 2)`×) compared to hippocampal differences (`r round(hc_int$SLOPE_RATIO, 2)`×) explains why HVR shows intermediate ratios (`r round(hvr_int$SLOPE_RATIO, 2)`×): HVR captures both processes, and the steeper male ventricular differences amplify the already steeper hippocampal differences. However, these interactions should be interpreted cautiously: @brzezinski-rittner_disentangling_2025 found that hippocampal age × sex interactions present in the full UK Biobank sample disappeared entirely in a TIV-matched subsample (N = 11,294), suggesting that some apparent sex-differential aging trajectories may be mathematical byproducts of initial brain size differences. Whether HVR age × sex interactions similarly reflect head-size confounding or genuine biological sex differences remains to be tested in matched longitudinal designs.

Several mechanisms may explain these sex-differential patterns. Cumulative estrogen exposure is protective: women with longer reproductive spans show younger brain ages (0.32 years per year of earlier menarche; @luders_case_2025), and earlier HRT initiation near menopause is associated with less evident brain aging in APOE ε4 carriers [@de_lange_womens_2020]. Estrogen promotes hippocampal neurogenesis, synaptic plasticity, mitochondrial function, and neuroinflammation reduction [@jett_endogenous_2022; @sundermann_female_2016; @diaz_brinton_healthy_2008], with the "critical period hypothesis" proposing that continuous exposure is necessary for neuroprotection. Menopause status may also moderate sex differences in tau pathology burden, with amyloid-related atrophy more strongly affecting women—an association likely mediated by higher tau accumulation [@bachmann_age-_2023].

Males also show earlier ventricular abnormality signatures, with ventricular enlargement accelerating in the seventies and preferential frontal horn involvement [@curra_ventricular_2019]. This earlier male ventricular expansion is consistent with the `r round(hvr_int$SLOPE_RATIO, 2)`× steeper male HVR decline observed here, suggesting that accelerated ventricular expansion—rather than differential hippocampal aging alone—drives the sex-differential HVR trajectory. The prognostic relevance of such trajectory differences is supported by evidence that steeper hippocampal volume loss during MCI predicts conversion to Alzheimer's dementia [@rajagopal_estimating_2024].

However, the female advantage is not universal: female APOE ε4 carriers show 2.3× faster hippocampal volume loss than male carriers in aMCI [@park_sex-specific_2025], indicating that genetic risk can modulate or reverse typical sex-differential patterns. Other large studies have found no significant sex differences in hippocampal age slopes [@fjell_minute_2009; @pintzka_marked_2015], underscoring that the presence and magnitude of age × sex interactions depends on sample size, age range, and head-size adjustment method. These cross-sectional patterns cannot establish true individual trajectories and may be confounded by cohort effects or differential survival. Longitudinal validation is essential.

## Normative Modeling (Analysis 4)

The sex-specific centile curves (@fig-normative-centiles) extend the growing literature on normative brain charts [@bethlehem_brain_2022; @dima_subcortical_2022; @ge_normative_2024] by providing the first sex-specific reference values for HVR alongside its component measures. While prior work has generated sex-stratified centiles for hippocampal volume [@dima_subcortical_2022; @ge_normative_2024] and global brain tissue volumes [@bethlehem_brain_2022], none have modeled the hippocampus-to-ventricle ratio. Our use of GAMLSS aligns with the methodological framework adopted by @bethlehem_brain_2022 and @dima_subcortical_2022, which allows flexible modeling of both non-linear age trends in the median and age-related changes in between-subject variability—a critical advantage over conventional models given the increasing variance observed at older ages, particularly for ventricular volumes.

Sex-specific norms are essential to avoid misclassification: a male at the 50th percentile for his sex would fall below the 25th percentile using female norms, potentially leading to false-positive concerns. This is consistent with recommendations from @williams_neuroanatomical_2021, who found that males exhibit greater variance in volumetric measures and concluded that "investigations of neuroanatomical deviance should always take sex into account," and with @ge_normative_2024, who developed sex-specific models for 150 brain measures because sex accounts for a considerable proportion of morphological variance. For HC, after residualization, sex differences in hippocampal centiles are minimal, consistent with the near-zero Cohen's d for residualized volumes, supporting the use of combined norms when proper ICV adjustment is applied. For LV, both sexes show larger ventricular volumes at older ages, but cross-sectional patterns and variability differ, with males showing greater ventricular variability at older ages—potentially reflecting greater heterogeneity in aging-related patterns.

The GAMLSS calibration plots (`r S("fig-gamlss-calibration")`) confirm that these centile estimates are well-calibrated, with observed proportions closely matching expected proportions in held-out test data. Individualized centile scores have demonstrated clinical utility in detecting case-control differences across neurological disorders and in quantitatively predicting diagnostic transition from MCI to Alzheimer's disease [@bethlehem_brain_2022; @dima_subcortical_2022]. Values below the 5th centile may warrant clinical follow-up, and longitudinal decline of more than one major centile band (e.g., 50th to 25th) may indicate accelerating age-related change.

## Brain-Cognition Associations (Analysis 5)

### Measurement Invariance

The successful establishment of strict measurement invariance for the cognitive bifactor model (`r S("tbl-measurement-invariance")`) is a prerequisite for the brain-cognition analyses. Without measurement invariance, apparent sex differences in brain-cognition paths could reflect measurement artifacts rather than true structural differences [@chen_sensitivity_2007; @putnick_measurement_2016]. The $\Delta$CFI $\leq$ 0.01 and $\Delta$RMSEA $\leq$ 0.015 at each invariance level [@cheung_evaluating_2002; @chen_sensitivity_2007] confirms that the same cognitive constructs (*g*, Memory~s~, Speed~s~) exist in both sexes, that factor loadings are equivalent (meaning the tests measure constructs with equal precision), that intercepts are equivalent (allowing valid comparison of factor means), and that residual variances are equivalent (confirming comparable reliability). This finding validates the use of the same factor structure for both sexes in the SEM analyses and supports the comparability of path coefficients across groups.

### Pooled and Sex-Stratified Paths

HVR predicted cognition as well as HC. Pooled model estimates with bootstrap CIs showed: HC → *g* $\beta$ = `r safe_round(hc_g$est.std, 3)` [`r safe_round(hc_g$ci.lower, 3)`, `r safe_round(hc_g$ci.upper, 3)`]; HVR → *g* $\beta$ = `r safe_round(hvr_g$est.std, 3)` [`r safe_round(hvr_g$ci.lower, 3)`, `r safe_round(hvr_g$ci.upper, 3)`]; HC~RES~ → *g* $\beta$ = `r safe_round(hc_res_g$est.std, 3)` [`r safe_round(hc_res_g$ci.lower, 3)`, `r safe_round(hc_res_g$ci.upper, 3)`]. HVR showed a numerically larger (though with overlapping CIs) association with general cognition despite requiring no head-size adjustment. These effect sizes are consistent with the small structural brain-cognition associations typically observed in large healthy samples: @ritchie_sex_2018 reported all brain-cognitive correlations below r = 0.20 in 5,216 UK Biobank participants, @wang_comparison_2024 found adjusted hippocampal → fluid intelligence coefficients of $\beta$ = 0.055 in 41,964 participants, and hippocampal volume-cognition associations in healthy aging cohorts are often non-significant at smaller sample sizes [@schoemaker_hippocampal-ventricle_2019; @bachmann_age-_2023]. The restricted cognitive range in healthy volunteer samples likely attenuates these associations [@bachmann_age-_2023].

Multigroup analyses indicated sex differences in brain-cognition associations. Residualized hippocampal volume (HC~RES~) was not significantly associated with cognition in males ($\beta$ = `r safe_round(hc_res_g_m$est.std, 3)`, 95% CI [`r safe_round(hc_res_g_m$ci.lower, 3)`, `r safe_round(hc_res_g_m$ci.upper, 3)`], p = `r safe_round(hc_res_g_m$pvalue, 3)`), while remaining significant in females ($\beta$ = `r safe_round(hc_res_g_f$est.std, 3)`, p < .001). This female-selective hippocampal-cognition association is consistent with @zheng_sex_2017, who found that hippocampal gray matter volume predicted associative memory in older women (r = 0.62) but not men, and with @jockwitz_differential_2024, who found that brain structure predicted cognitive profiles more strongly in females than males. Raw HC showed significant associations in both sexes but with a female advantage ($\beta$ = `r safe_round(hc_g_f$est.std, 3)` vs. `r safe_round(hc_g_m$est.std, 3)`). However, @ritchie_sex_2018 found no significant sex differences in brain-cognition correlations for verbal-numerical reasoning and reaction time, suggesting that sex-differential associations may be domain-specific.

HVR showed the opposite pattern: males exhibited *stronger* HVR-cognition associations ($\beta$ = `r safe_round(hvr_g_m$est.std, 3)`) than females ($\beta$ = `r safe_round(hvr_g_f$est.std, 3)`), though both reached significance. HVR captures the hippocampus-to-ventricle balance, and the male-stronger association suggests that this ratio may be more cognitively relevant in males, perhaps because males show steeper cross-sectional HVR differences (`r round(hvr_int$SLOPE_RATIO, 2)`× female rate) making individual differences more consequential.

For processing speed (Speed~s~), males showed strong HC and HC~RES~ associations ($\beta$ = `r safe_round(hc_speed_m$est.std, 3)` and `r safe_round(hc_res_speed_m$est.std, 3)`) while females showed none—suggesting sex-differential neural substrates for processing speed.

### Implications for the HVR Thesis

These sex-differential patterns support HVR in several ways. Unlike HC~RES~, which loses predictive validity in males for general cognition, HVR maintains significant associations in both sexes. @schoemaker_hippocampal-ventricle_2019 similarly found that HVR explained more variance in memory scores than absolute hippocampal volume (40% vs. 32% in their first cohort), with HC failing to reach significance as a predictor in models where HVR was significant. @fernandez-lozano_enhanced_2025 extended this finding to clinical populations, showing that HVR associations with cognition were significant across healthy, MCI, and AD groups and were comparable in magnitude to hippocampal volume. Given that males show steeper cross-sectional HVR differences with age and stronger HVR-cognition associations, HVR may be useful for tracking brain aging in males. The differential patterns between HC~RES~ and HVR across sexes indicate HVR is not simply "residualized HC by another means" but captures distinct information related to the hippocampus-ventricle coupling. The association with general cognition (*g*) rather than domain-specific factors may reflect that lateral ventricular volume acts as a marker for widespread tissue loss beyond the medial temporal lobe: @bachmann_age-_2023 found that ventricular volume was exclusively associated with executive function rather than episodic memory, suggesting that HVR's ventricular component captures cognitive-relevant variance beyond hippocampal integrity.

### Model Stability and Non-Linear Age Effects

SEM analyses indicated that raw hippocampal volume (HC) required a linear-age-only specification because the combination of HC, ICV, and age² produced multicollinearity (information matrix not invertible). In contrast, both HVR and residualized HC (HC~RES~) successfully incorporated quadratic age terms, which were significant predictors of both brain structure and cognition (all p < .001). Quadratic age effects were: HVR ~ age² $\beta$ = `r safe_round(hvr_age_sq$est.std, 3)`, indicating steeper HVR differences at older ages; HC~RES~ ~ age² $\beta$ = `r safe_round(hc_res_age_sq$est.std, 3)`; g ~ age² $\beta$ = `r safe_round(g_age_sq_hvr$est.std, 3)`; and Speed~s~ ~ age² $\beta$ = `r safe_round(speed_age_sq_hvr$est.std, 3)`, with processing speed showing the expected age vulnerability.

The stronger quadratic effect for HVR ($\beta$ = `r safe_round(hvr_age_sq$est.std, 3)`) compared to HC~RES~ ($\beta$ = `r safe_round(hc_res_age_sq$est.std, 3)`) reflects that HVR captures both hippocampal and ventricular age-related differences, which both show steeper cross-sectional patterns at older ages. HVR detects accelerating age-related patterns. Researchers using raw hippocampal volumes in complex multivariate models may encounter convergence issues or require simpler covariate structures, potentially missing non-linear relationships. Self-normalizing measures like HVR, or statistically residualized volumes, allow for non-linear modeling.

## Sensitivity Analyses (Analysis 6)

### Site Effects

The low site ICCs (`r S("tbl-site-adjusted")`, all < 0.05) indicate that scanner site contributed minimally to variance in brain measures. This is consistent with UK Biobank's harmonized acquisition protocol across sites [@alfaro-almagro_image_2018]. The stability of effect sizes after site adjustment confirms that our findings are not driven by site-specific biases.

Different analytical contexts employed appropriate site-handling methods. Normative GAMLSS models used random intercepts for site to allow site-specific baseline levels while pooling information across sites. Sex difference analyses used linear mixed-effects models with random site intercepts. Pooled SEM analyses used cluster-robust standard errors grouped by site rather than random effects, as the FIML estimator in lavaan does not support random effects. Multi-group SEM analyses used robust maximum likelihood estimation without site clustering, as lavaan does not support the combination of multi-group analysis with clustered standard errors.

For HVR specifically, site effects could theoretically differentially affect hippocampus and ventricle measurements. The low site ICC for HVR suggests that the ratio construction does not amplify site-related noise.

### Hemisphere Comparison

For individual volumes, sex differences were symmetrical across hemispheres: left and right hippocampal effect sizes were nearly identical (d = −0.14 and −0.16, overlapping CIs), as were ventricular effect sizes (d = 0.48 bilaterally). This bilateral symmetry is consistent with @fjell_minute_2009, who found a non-significant hemisphere × sex × structure interaction (p = 0.97) across 13 brain volumes including hippocampus and lateral ventricles, and with @ritchie_sex_2018, who reported nearly identical left (d = −0.55) and right (d = −0.54) hippocampal sex differences in the UK Biobank.

However, HVR showed a lateralized pattern, with a significant female advantage in the right hemisphere (d = 0.05) but not the left (d = −0.01). Because HVR combines two bilaterally symmetrical structures into a ratio, this lateralization likely reflects subtle asymmetries in how hippocampal and ventricular volumes interact within each hemisphere rather than lateralized sex differences in either structure alone. @ruigrok_meta-analysis_2014 noted that female volume increases tend to be more pronounced in right limbic structures, which could contribute to a right-lateralized HVR advantage. However, automated segmentation algorithms show hemispheric biases—higher sensitivity for the right hippocampus but better ventricular segmentation on the left [@fernandez-lozano_enhanced_2025]—which could produce artifactual laterality in a derived ratio measure. The small magnitude of this lateralization (d = 0.05 vs. −0.01) and its potential methodological origins suggest that bilateral HVR remains the most robust measure for clinical application.

### Inclusion of Psychiatric Diagnoses

Including participants with psychiatric diagnoses (F-codes) resulted in minimal change in effect sizes ($\Delta$d < 0.02; `r S("tbl-sensitivity")`, `r S("fig-sensitivity")`), with all findings retaining the same direction and significance. This stability is expected given the low prevalence of psychiatric diagnoses in the UK Biobank, which reflects the well-documented healthy volunteer selection bias of the cohort [@ritchie_sex_2018; @fry_comparison_2017]. While reduced hippocampal volume is a shared feature across major psychiatric conditions [@brosch_reduced_2022; @schmaal_subcortical_2016], the low base rates in this sample mean that inclusion or exclusion of F-code participants lacks the statistical weight to meaningfully shift population-level estimates. Nonetheless, establishing robustness to psychiatric exclusion criteria is important because the brain regions showing the largest sex differences (hippocampus, amygdala) overlap with those implicated in sex-biased psychiatric conditions, creating potential sex-by-disorder interactions that could confound normative sex difference estimates [@ruigrok_meta-analysis_2014; @ritchie_sex_2018].

## Limitations

This study has limitations related to design, sample, and methodology. The cross-sectional design cannot distinguish age effects from cohort effects or establish individual change trajectories, and HVR conflates two structures, precluding separation of hippocampal and ventricular contributions to observed differences.

Sample characteristics impose additional constraints. UK Biobank participants are healthier and more educated than the general population [@fry_comparison_2017], limiting generalizability to clinical populations. Some cognitive tests had incomplete data, and complete-case analyses may introduce selection bias. APOE genotype was unavailable, precluding examination of gene × sex interactions relevant to dementia risk.

Several methodological considerations warrant mention. The study did not include the power-corrected proportions allometric adjustment method proposed by @sanchis-segura_effects_2020. Imaging site clustering was not modeled in multigroup SEM because lavaan does not support the combination of multigroup analysis with clustered standard errors [@rosseel_lavaan_2012]. However, with only 4 imaging sites across approximately 27,000 participants (average cluster size >6,000), the design effect is minimal, and any residual site effects would inflate standard errors equally across sexes, making sex comparisons conservative rather than biased.

The bifactor model's memory-specific factor (Memory~s~) could not be reliably estimated in either sex (females: estimation failure; males: all loadings non-significant, p > .90). This "factor collapse" is common in bifactor models, occurring in approximately 60% of published applications [@eid_anomalous_2017; @burns_application_2020], indicating that variance in memory test performance was fully captured by the general cognitive factor (*g*). Brain-cognition paths involving Memory~s~ were therefore not interpretable and are not reported, though the processing speed factor (Speed~s~) was well-identified in both sexes.

Ratios can exhibit non-normality, heteroscedasticity, and complex interpretation when numerator and denominator correlate differently with outcomes; while HVR showed approximately normal distributions in this sample, these properties should be verified in other populations. All brain-cognition associations were small ($\beta$ = 0.02–0.04), explaining less than 1% of variance in cognitive performance. While statistically significant given the large sample, the practical and clinical significance of these associations remains to be established.

## Future Directions

The current findings motivate four lines of follow-up work. First, the UK Biobank repeat imaging subsample enables longitudinal validation of whether HVR predicts individual-level cognitive decline. Second, clinical validation in AD/MCI cohorts would assess HVR's diagnostic utility compared to clinical populations studied previously [@fernandez-lozano_enhanced_2025]. Third, longitudinal data could separate hippocampal and ventricular contributions to HVR changes, which the cross-sectional ratio conflates. Fourth, examining whether HVR × sex × APOE genotype predicts dementia conversion would address the gene × sex interactions that this study could not explore.

## Conclusions

In `r format(n_primary, big.mark=",")` UK Biobank participants (`r pct_female`% female), HVR demonstrated substantial head-size independence (r~ICV~ = `r round(r_hvr_icv, 2)`; `r icv_reduction_pct`% reduction vs. raw HC), a consistent female advantage replicated in ICV-matched samples (d = `r round(matched_hvr$ESTIMATE_MTCH, 2)`), and comparable brain-cognition associations to hippocampal volume (HVR → *g*: $\beta$ = `r safe_round(hvr_g$est.std, 3)` vs. HC → *g*: $\beta$ = `r safe_round(hc_g$est.std, 3)`, overlapping CIs). In contrast, hippocampal sex differences reversed direction across adjustment methods (d range: `r round(abs(hc_d_range), 1)` SD), confirming their methodological determination. HVR maintained predictive validity in both sexes, while residualized HC lost significance in males (p = `r safe_round(hc_res_g_m$pvalue, 3)`), and supported more complex SEM specifications that raw HC could not.

Males showed `r round(hvr_int$SLOPE_RATIO, 2)`× steeper age-related HVR differences (p = `r format(hvr_int$INTERACTION_P, scientific=TRUE, digits=2)`), consistent with earlier ventricular expansion signatures in males [@curra_ventricular_2019] and evidence that hippocampal trajectory slopes predict dementia conversion [@rajagopal_estimating_2024]. Sex-specific normative centile curves are provided for research application and potential clinical use, pending validation in clinical populations.

# Author Contributions {.unnumbered}

S.F-L. conceived the study, developed the analysis pipeline, performed all analyses, and wrote the manuscript. D.L.C. supervised the project, provided critical review, and approved the final manuscript.

# Acknowledgments {.unnumbered}

This research has been conducted using the UK Biobank Resource under Application Number 45551. This investigation was supported in part by an award from the International Progressive MS Alliance (award reference number PA-1412-02420), a grant from the Canadian Institutes of Health Research, and a donation from the Famille Louise & Andre Charron.

# Conflict of Interest Statement {.unnumbered}

The authors declare no conflicts of interest.

# Data Availability Statement {.unnumbered}

This research was conducted using UK Biobank data (Application 45551). Data are available to approved researchers through the UK Biobank Access Management System (https://www.ukbiobank.ac.uk/). Analysis code is available at https://github.com/soffiafdz/hvr-sex-differences.

# Ethics Approval Statement {.unnumbered}

UK Biobank received ethical approval from the North West Multi-centre Research Ethics Committee (REC reference: 16/NW/0274). All participants provided written informed consent.

# References {.unnumbered}

::: {#refs}
:::
