---
title: "Hippocampal-to-Ventricle Ratio as a Head-Size-Independent Biomarker: Sex Differences and Cognitive Associations in 27,680 UK Biobank Participants"
running-title: "HVR Sex Differences in UK Biobank"
keywords:
  - Hippocampus
  - Lateral Ventricles
  - Sex Characteristics
  - Magnetic Resonance Imaging
  - Cognitive Aging
  - Biomarkers
  - Reference Values
key-points:
  - Hippocampal sex differences reverse direction depending on head-size adjustment method, spanning 1.5 SD across analytical choices
  - The hippocampal-to-ventricle ratio (HVR) provides consistent sex difference estimates without requiring head-size correction and predicts cognition comparably to adjusted hippocampal volume
  - Males show steeper cross-sectional HVR decline with age; sex-specific normative centile curves are provided for clinical application
author:
  - name: Sofia Fernandez-Lozano
    affiliations:
      - ref: bic
      - ref: ipn
  - name: D. Louis Collins
    affiliations:
      - ref: bic
      - ref: ipn
      - ref: bbme
    email: louis.collins@mcgill.ca
    corresponding: true
affiliations:
  - id: bic
    name: McConnell Brain Imaging Centre
    department: Montreal Neurological Institute
    city: Montreal
    state: Quebec
    country: Canada
  - id: ipn
    name: McGill University
    department: Department of Neurology and Neurosurgery
    city: Montreal
    state: Quebec
    country: Canada
  - id: bbme
    name: McGill University
    department: Department of Biomedical Engineering
    city: Montreal
    state: Quebec
    country: Canada
date: today
abstract: |
  Hippocampal volume is a key biomarker for Alzheimer's disease and brain aging, yet reported sex differences vary substantially depending on head-size adjustment methods. We examined whether the hippocampal-to-ventricle ratio (HVR), a self-normalizing measure, provides more consistent sex difference estimates than conventional adjustment approaches. Using a subset of UK Biobank structural MRI data (N = 27,680; 56% female; ages 45--82), we compared sex differences across four adjustment methods (unadjusted, proportions, stereotaxic, residualized) and in samples matched on age and on intracranial volume (ICV). Hippocampal sex differences reversed direction across methods, ranging from $d = -0.89$ (males larger, unadjusted) to $d = 0.58$ (females larger, proportions), with a range of 1.5 standard deviations across analytical choices. In contrast, HVR showed a consistent female advantage ($d = 0.52$) that persisted in ICV-matched samples ($d = 0.19$), confirming this effect is not a head-size artifact. Males exhibited steeper cross-sectional age-related HVR decline ($1.7\times$ female rate, $p < 10^{-50}$), consistent with males showing ventricular expansion at younger ages. Structural equation modeling revealed that HVR predicted general cognitive ability comparably to hippocampal volume ($\beta = 0.04$, overlapping CIs), and unlike residualized hippocampal volume, HVR maintained significant brain-cognition associations in both sexes. We provide sex-specific normative centile curves for clinical application. These findings indicate that apparent hippocampal sex differences largely reflect methodological choices rather than biology, while HVR captures consistent morphological variation related to brain aging that may have clinical utility.
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    include-in-header:
      text: |
        <style>
        .gt_table, .gt_table th, .gt_table td {
          font-size: 14px !important;
        }
        table, th, td {
          font-size: 14px !important;
        }
        </style>
  pdf:
    toc: false
    number-sections: true
    documentclass: article
    geometry:
      - margin=1in
    latex-max-runs: 20
    pdf-engine: xelatex
    keep-tex: true
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage{array}
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{amsmath}
        \usepackage{amssymb}
        \usepackage{textcomp}
        \usepackage{fontspec}
        \setmainfont{Latin Modern Roman}
        \usepackage{newunicodechar}
        \newunicodechar{ε}{\ensuremath{\varepsilon}}
        \usepackage{authblk}
        \renewcommand\Authfont{\normalsize}
        \renewcommand\Affilfont{\small\itshape}
        % Force tables to appear exactly where placed (prevent float drift)
        \usepackage{tikz}
        \usetikzlibrary{shapes.geometric, arrows.meta, positioning}
        \usepackage{float}
        \floatplacement{table}{H}
        % Center longtables
        \setlength{\LTleft}{\fill}
        \setlength{\LTright}{\fill}
  rtf:
    toc: false
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
  eval: true
bibliography: references.bib
csl: human-brain-mapping.csl
---

```{r}
#| label: setup
#| include: false

# Activate renv (quarto runs R with --vanilla, skipping .Rprofile)
renv::load(file.path(getwd(), ".."))

library(here)
library(data.table)
library(gt)
here::i_am("reports-src/manuscript.qmd")
source(here("reports-src/_common.R"))

# Load pre-computed manuscript environment (from R/12_manuscript_objects.R)
ms <- read_rds_safe(here("data/derivatives/manuscript_env.rds"), "manuscript environment")
list2env(ms$scalars, envir = environment())
list2env(ms$data, envir = environment())
list2env(ms$tables, envir = environment())

```

# Introduction

The hippocampus (HC) is a brain structure used as an early biomarker for Alzheimer's disease (AD) and brain aging [@jack_prediction_1999; @jack_hypothetical_2010; @raz_regional_2005]. Sex differences in brain structures have been documented, but their interpretation is controversial [@ritchie_sex_2018; @ruigrok_meta-analysis_2014], largely because findings depend on whether and how head-size adjustments are applied [@sanchis-segura_effects_2020; @nordenskjold_intracranial_2015]. Males have approximately 12% larger intracranial volume (ICV) than females [@ruigrok_meta-analysis_2014], and after head-size correction, sex classification from brain features drops from >80% to approximately 60% [@sanchis-segura_effects_2020]. This methodological sensitivity suggests some reported sex differences may be artifacts of adjustment methods.

Hippocampal atrophy rates predict conversion from mild cognitive impairment to dementia [@rajagopal_estimating_2024], making accurate volume assessment clinically consequential. However, males and females show different rates of age-related brain volume change, with steeper subcortical volume declines in males across multiple structures [@ritchie_sex_2018]. These sex-differential trajectories mean that normative reference values must be sex-stratified to avoid misclassifying normal variation as pathological; recent large-scale brain charting efforts have adopted sex-specific centile modeling for this reason [@bethlehem_brain_2022; @ge_normative_2024].

Multiple adjustment methods exist, each with different assumptions: proportions (V/ICV) assumes perfect allometric scaling, residualization (V $-$ $\beta \times$ ICV) removes linear ICV association, and stereotaxic normalization applies spatial transformation to a template. These methods can yield contradictory conclusions about the same brain regions [@wang_comparison_2024]. For hippocampus, reported sex differences range from males larger (unadjusted; @ruigrok_meta-analysis_2014) to females larger (proportions; @sanchis-segura_effects_2020) to no difference (residualized; @obrien_statistical_2011). The proportions method can produce biologically implausible results, while residualization yields more consistent estimates [@wang_comparison_2024]. No consensus exists on the correct method, and conclusions depend on analytical choices.

The hippocampal-to-ventricle ratio (HVR), defined as HC / (HC + LV) where HC denotes total hippocampal volume and LV denotes the volume of the temporal horns of the lateral ventricles, offers a potential solution. Both structures scale with head size, making their ratio self-normalizing. HVR captures the relationship between hippocampal and ventricular volumes, which show opposite patterns with age—smaller hippocampus and larger ventricles. Schoemaker and colleagues introduced HVR and showed that it provides stronger associations with age and memory than standard hippocampal volume in healthy aging [@schoemaker_hippocampal-ventricle_2019]. Subsequent work shows HVR provides higher effect sizes for distinguishing AD from controls compared to HC alone [@fernandez-lozano_enhanced_2025]. However, properties of HVR in healthy populations and across adjustment methods remain unexplored.

Age-related ventricular expansion reflects concurrent loss of surrounding gray and white matter tissue [@madsen_mapping_2015], and this expansion occurs earlier in males, with abnormal enlargement appearing in the seventies and preferentially affecting the frontal horns [@curra_ventricular_2019]. These sex-differential ventricular trajectories suggest that a ratio incorporating both hippocampal and ventricular volumes may capture sex differences in brain aging that individual volumes miss. Furthermore, brain structure–cognition associations themselves may differ by sex: grey matter volumes predict cognitive profiles with different strength and spatial specificity in older males versus females [@jockwitz_differential_2024], and hippocampal volume predicts associative memory in older women but not men [@zheng_sex_2017]. Whether HVR shows similar sex-differential cognitive associations remains unknown.

We addressed three questions: (1) Do hippocampal sex differences vary across head-size adjustment methods while HVR remains consistent? (2) Does HVR show comparable cognitive associations to conventionally adjusted hippocampal volumes? (3) Do cross-sectional age patterns differ by sex, and do these patterns differ between HC and HVR depending on normalization method?

---

# Materials and Methods

## Data Source and Participants

Data were obtained from UK Biobank (Application 45551) [@sudlow_uk_2015; @miller_multimodal_2016]. The initial imaging sample comprised approximately `r format(n_initial_imaging, big.mark=",")` participants with T1-weighted structural MRI. After exclusions (detailed below), the final analytic sample included N = `r format(n_primary, big.mark=",")` participants aged `r age_min` to `r age_max` years. With this sample size, we had >99% power to detect small effects (d = 0.10) and >80% power to detect very small effects (d = 0.05) for two-group comparisons at $\alpha$ = .05.

Three additional subsamples were defined for specific analyses (@fig-prisma; @tbl-demographics). The **England subsample** (N = `r format(n_england, big.mark=",")`) comprised participants assessed at UK Biobank centres in England and was used for SEM analyses requiring the Index of Multiple Deprivation (IMDP), which is not comparable across UK countries. This subsample provides adequate power for detecting standardized path coefficients $\geq$ 0.02. The **ICV-matched subsample** (N = `r format(n_matched_pairs * 2, big.mark=",")`) comprised female--male pairs matched on age ($\pm 1$ year) and ICV ($\pm 25$ cc) drawn from the primary sample, and was used to isolate sex effects from head-size confounding (Analysis 2). The **longitudinal subsample** (N = 2,935) comprised participants from the primary sample who also had a follow-up imaging session (ses-3), with a mean intersession interval of 2.7 years (SD = 1.1; range: 1.5--5.0). This subsample was used as a sensitivity check for normative model calibration (Analysis 4).

### Exclusion Criteria

Participants were excluded if they had any of the following ICD-10 diagnoses recorded in linked hospital records: F00-F99 (mental and behavioral disorders including dementia, mood disorders, schizophrenia, and anxiety disorders, which are associated with hippocampal volume alterations [@schmaal_subcortical_2016; @cao_volumes_2023; @brosch_reduced_2022]); G00-G99 (neurological diseases including Parkinson's disease, multiple sclerosis, epilepsy, and cerebrovascular disease, which directly affect brain structure [@van_erp_subcortical_2016; @xu_longitudinal_2020; @koenig_hippocampal_2014; @tai_detection_2025]); and Q00-Q07 (congenital CNS malformations affecting normal brain development). This exclusion strategy selected a sample of cognitively healthy adults without conditions known to affect hippocampal morphometry. The full exclusion flow is shown in @fig-prisma.

::: {#fig-prisma}

```{=latex}
\centering
\begin{tikzpicture}[
  node distance=0.8cm and 1.6cm,
  every node/.style={font=\small},
  process/.style={rectangle, draw=black!70, thick, text width=4.2cm,
                  minimum height=0.9cm, align=center, rounded corners=3pt},
  decision/.style={rectangle, draw=black!70, thick, text width=3.2cm,
                   minimum height=0.9cm, align=center, rounded corners=3pt,
                   fill=black!5},
  excluded/.style={rectangle, draw=black!70, thick, text width=3.2cm,
                   minimum height=0.9cm, align=center, rounded corners=3pt,
                   fill=black!8, text=black!70},
  sample/.style={rectangle, draw=black!70, very thick, text width=4.2cm,
                 minimum height=0.9cm, align=center, rounded corners=3pt},
  subsample/.style={rectangle, draw=black!70, thick, text width=3.6cm,
                    minimum height=0.9cm, align=center, rounded corners=3pt,
                    fill=black!3, dashed},
  arrow/.style={-{Stealth[length=5pt]}, thick, black!70}
]
  % Main flow (top to bottom)
  \node[process] (initial)
    {UK Biobank Imaging Cohort\\[2pt]\textbf{N\,=\,47,398}};

  \node[decision, below=of initial] (qc)
    {Quality control filter};

  \node[process, below=of qc] (passed)
    {Passed all processing\\[2pt]\textbf{n\,=\,38,567}};

  \node[decision, below=of passed] (gq)
    {ICD-10 exclusion:\\G00--G99, Q00--Q07};

  \node[sample, below=of gq] (sensitivity)
    {Sensitivity sample\\[2pt]\textbf{N\,=\,30,032}};

  \node[decision, below=of sensitivity] (f)
    {ICD-10 exclusion:\\F00--F99};

  \node[sample, below=of f] (primary)
    {\textbf{Primary sample}\\[2pt]\textbf{N\,=\,27,680}};

  % Exclusion boxes (right side)
  \node[excluded, right=of qc] (exc1)
    {Excluded\\[2pt]n\,=\,8,831};

  \node[excluded, right=of gq] (exc2)
    {Excluded\\[2pt]n\,=\,8,535};

  \node[excluded, right=of f] (exc3)
    {Excluded\\[2pt]n\,=\,2,352};

  % Subsamples (below primary, three across)
  \node[subsample, below=1.2cm of primary, xshift=-4.2cm] (england)
    {England only\\[2pt]\textit{N\,=\,25,172}\\[-1pt]{\scriptsize(SEM analyses)}};

  \node[subsample, below=1.2cm of primary] (matched)
    {ICV-matched\\[2pt]\textit{N\,=\,9,500}\\[-1pt]{\scriptsize(Sex differences)}};

  \node[subsample, below=1.2cm of primary, xshift=4.2cm] (longitudinal)
    {Longitudinal\\[2pt]\textit{N\,=\,2,935}\\[-1pt]{\scriptsize(Normative calibration)}};

  % Arrows
  \draw[arrow] (initial) -- (qc);
  \draw[arrow] (qc) -- (passed);
  \draw[arrow] (qc) -- (exc1);
  \draw[arrow] (passed) -- (gq);
  \draw[arrow] (gq) -- (sensitivity);
  \draw[arrow] (gq) -- (exc2);
  \draw[arrow] (sensitivity) -- (f);
  \draw[arrow] (f) -- (primary);
  \draw[arrow] (f) -- (exc3);
  \draw[arrow, dashed] (primary.south) -- ++(0,-0.35cm) -| (england.north);
  \draw[arrow, dashed] (primary.south) -- (matched.north);
  \draw[arrow, dashed] (primary.south) -- ++(0,-0.35cm) -| (longitudinal.north);
\end{tikzpicture}
```

**Participant selection flow.** Dashed boxes indicate analytic subsamples derived from the primary sample.
:::

### Quality Control

Image segmentations underwent quality control in three stages: (1) automated quality scores from deep learning-based assessments, requiring DARQ score $\geq$ 0.25 [@fonov_darq_2022] and AssemblyNet rating of "A" (highest quality grade) [@coupe_assemblynet_2020], (2) statistical outlier detection excluding participants with age- and head-size-adjusted volumes exceeding $\pm 3$ SD from the sample mean, and (3) visual inspection of flagged cases by trained raters for evident segmentation failures, motion artifacts, or anatomical anomalies.

For participants with multiple imaging sessions, QC filtering was applied first, then one session per participant was selected for cross-sectional analyses. When a participant's baseline session (ses-2) failed QC but the follow-up session (ses-3) passed, the follow-up session was retained in the cross-sectional sample. Longitudinal analyses required both sessions to pass QC.

## Brain Measures

### Image Acquisition

Images were acquired on Siemens Skyra 3T scanners at three UK sites (Cheadle, Newcastle, Reading) using a T1-weighted MPRAGE sequence at 1mm isotropic resolution. Full protocol details are provided in @alfaro-almagro_image_2018.

### Segmentation

Hippocampal and ventricular volumes were segmented using an in-house convolutional neural network trained on manually labeled data. Validation against manual segmentations showed Dice $\kappa$ = 0.94 and Pearson r = 0.99 [@fernandez-lozano_enhanced_2025]. Outputs included bilateral hippocampal volumes (left, right, total) and lateral ventricular volumes (left, right, total). Intracranial volume (ICV) was obtained from AssemblyNet [@coupe_assemblynet_2020], a large-ensemble convolutional neural network for whole-brain MRI segmentation.

### Head-Size Adjustment Methods

Four adjustment approaches were applied to hippocampal and ventricular volumes [@sanchis-segura_effects_2020; @sanfilipo_correction_2004]:

1. **Unadjusted**: raw volumes in native space (mL), which preserve natural variation but are confounded by overall head size.

2. **Proportions**: volume expressed as a proportion of ICV, which assumes isometric (1:1) scaling and may over- or under-correct depending on the true allometric relationship [@de_jong_allometric_2017; @williams_neuroanatomical_2021]:

$$V_{\text{PRP}} = \frac{V}{\text{ICV}} \times 1000$$ {#eq-prp}

3. **Stereotaxic**: volumes in standard template space after nonlinear registration, which corrects for global size and shape differences:

$$V_{\text{STX}} = V \times |J|$$ {#eq-stx}

where $J$ is the Jacobian determinant of the nonlinear transformation.

4. **Residualized**: linear regression residuals after removing ICV association, with regression coefficients estimated from the pooled sample [@sanchis-segura_effects_2020; @obrien_statistical_2011]:

$$V_{\text{RES}} = V - \hat{\beta} \times (\text{ICV} - \overline{\text{ICV}})$$ {#eq-res}

Residualization eliminates ICV-related variance and produces more replicable sex differences compared to proportional methods.

### HVR Calculation

The hippocampal-to-ventricle ratio (HVR) was computed as:

$$\text{HVR} = \frac{\text{HC}_{\text{total}}}{\text{HC}_{\text{total}} + \text{LV}_{\text{total}}}$$ {#eq-hvr}

where $\text{HC}_{\text{total}}$ and $\text{LV}_{\text{total}}$ denote bilateral sums (left + right) of hippocampal and lateral ventricular volumes, respectively. HVR is bounded between 0 and 1, with higher values indicating larger hippocampus relative to ventricles. Because HVR is a ratio of two head-size-dependent structures, no further head-size adjustment was applied.

## Cognitive Assessment

### Test Battery

Cognitive data were collected at the imaging visit using the UK Biobank touchscreen-based cognitive battery [@sudlow_uk_2015], comprising 12 indicators spanning memory, processing speed, executive function, and reasoning domains (`r S("tbl-cog-battery")`):

1. **Pair matching -- average time** (mean response time; inverse-coded)
2. **Pair matching -- average errors** (mean incorrect responses; inverse-coded)
3. **Numeric memory** (maximum digits recalled)
4. **Prospective memory** (recall timing: 0 = incorrect, 1 = second attempt, 2 = first attempt)
5. **Fluid intelligence** (total score)
6. **Matrix pattern recognition** (puzzles solved)
7. **Trail making -- numeric** (completion time; inverse-coded)
8. **Trail making -- alphanumeric** (completion time; inverse-coded)
9. **Tower rearranging** (correct puzzles)
10. **Reaction time** (mean time to identify matches; inverse-coded)
11. **Symbol digit substitution -- correct** (correct responses)
12. **Symbol digit substitution -- attempts** (total attempts)

Missing cognitive data were handled using full information maximum likelihood (FIML) estimation, which uses all available data under a missing-at-random assumption.

### Factor Structure: Bifactor CFA

Cognitive indicators were modeled using a bifactor confirmatory factor analysis (CFA) [@sudlow_uk_2015] with three latent factors. The general factor (*g*) loaded on all 12 indicators. The memory-specific factor (Memory~s~, orthogonal to *g*) loaded on pair matching, numeric memory, and prospective memory. The processing speed-specific factor (Speed~s~, orthogonal to *g*) loaded on reaction time, trail making numeric, and symbol digit substitution correct. Factor scales were set using the marker variable method (first loading fixed to 1.0 for each factor). All latent factors were constrained to be orthogonal, consistent with bifactor model assumptions. Cross-loadings were fixed to zero. Residual covariances were specified for indicator pairs sharing method variance within the same test (pair matching time with errors; symbol digit correct with attempts; trail making numeric with alphanumeric).

Models were estimated using maximum likelihood with robust standard errors (MLR) in lavaan [@rosseel_lavaan_2012].

## Statistical Analyses

### Analysis 1: HVR Self-Normalizing Properties

To evaluate whether HVR reduces head-size confounding, we computed Pearson correlations between ICV and each brain measure (HC, LV, HVR) as well as between ICV and each head-size adjusted HC volume (proportions, stereotaxic, residualized). We expected a self-normalizing measure to show reduced correlation with ICV compared to unadjusted volumes.

### Analysis 2: Sex Differences in Brain Measures

Sex differences were quantified using Cohen's d with 95% confidence intervals [@cohen_statistical_2013], with positive d indicating females > males by our coding convention. Effect sizes were computed for all ROI x adjustment method combinations in the full sample (N = `r format(n_primary, big.mark=",")`).

To isolate sex effects from head-size confounding, we created a matched subsample where females and males were paired on age ($\pm 1$ year) and ICV ($\pm 25$ cc, approximately 0.3 SD). Matching used the MatchIt package (v4.7.2; @ho_matchit_2011) with optimal 1:1 matching without replacement (`method = "optimal"`, `distance = "mahalanobis"`), with calipers enforcing tolerance bounds. This yielded N = `r format(n_matched_pairs, big.mark=",")` matched pairs (total N = `r format(n_matched_pairs * 2, big.mark=",")`). We reasoned that if sex differences in HC persist after matching on ICV, they cannot be attributed solely to head-size confounding, and if HVR shows similar effects in full and matched samples, this indicates HVR is independent of head size.

### Analysis 3: Age x Sex Interactions

Linear regression models tested whether cross-sectional age-related patterns differed by sex:

$$Y_i = \beta_0 + \beta_{\text{Age}} \cdot \text{Age}_i + \beta_{\text{Sex}} \cdot \text{Sex}_i + \beta_{\text{Age} \times \text{Sex}} \cdot \text{Age}_i \times \text{Sex}_i + \varepsilon_i$$ {#eq-age-sex}

where $Y$ is the brain measure, Sex is coded as a binary indicator (Female = reference), and $\varepsilon_i \sim \mathcal{N}(0, \sigma^2)$. A significant $\beta_{\text{Age} \times \text{Sex}}$ indicates different age slopes between sexes. Sex-specific slopes were derived as $\hat{\beta}_{\text{Age}}$ for females and $\hat{\beta}_{\text{Age}} + \hat{\beta}_{\text{Age} \times \text{Sex}}$ for males, and their ratio quantified the relative rate of cross-sectional age-related differences. Separate models were fit for HC~RES~ and LV~RES~ (both residualized against ICV; @eq-res) and HVR (self-normalizing; @eq-hvr). Scanner site was not included as a random effect given the low site ICCs (< 0.05; Analysis 6), indicating minimal site contribution to variance. While age trajectories may be non-linear (as captured by the GAMLSS normative models below), linear interaction models were used here to provide interpretable summary statistics for sex-differential aging rates.

### Analysis 4: Normative Modeling

Sex-specific normative centile curves were estimated using Generalized Additive Models for Location, Scale, and Shape (GAMLSS) [@rigby_generalized_2005; @bethlehem_brain_2022]. For HC and LV (unadjusted):

$$\mu = f_\mu(\text{Age}) + \beta_{\text{ICV}} \cdot \text{ICV} + \beta_{\text{Edu}} \cdot \text{Education} + u_{\text{site}}$$ {#eq-gamlss-hc}

For HVR (self-normalizing, no ICV term needed):

$$\mu = f_\mu(\text{Age}) + \beta_{\text{Edu}} \cdot \text{Education} + u_{\text{site}}$$ {#eq-gamlss-hvr}

where $f_\mu(\cdot)$ denotes a cubic spline smooth for the location parameter and $u_{\text{site}}$ denotes random intercepts for scanner site. The scale parameter was also modeled as a function of age, $\sigma = f_\sigma(\text{Age})$, allowing variance to change across the age range. Model families (Gaussian and Box-Cox Cole and Green; BCCG) were compared using AIC, with BIC as secondary criterion. The BCCG distribution was used when flexible modeling of location ($\mu$), scale ($\sigma$), and skewness ($\nu$) was warranted; shape parameters beyond $\sigma$ were held constant to maintain parsimony. Centile curves (5th, 10th, 25th, 50th, 75th, 90th, 95th) were generated for ages 45--82 by sex.

Data were split into 80% training and 20% test sets, stratified by sex and 5-year age bins. Distribution families were selected on training data and evaluated on the held-out test set by comparing observed vs. expected proportions across centiles. Final models were then fit on the complete cross-sectional sample using the selected distribution family. Model calibration was additionally assessed using the longitudinal subsample (N = 2,935; see Data Source and Participants). Because this subsample comprises the same participants as the primary cross-sectional sample with an additional follow-up session, it tests temporal stability rather than independent validation. Test-retest reliability was quantified using Pearson correlation and intraclass correlation coefficient (ICC, two-way random, absolute agreement).

Two sets of GAMLSS models were generated: (1) site-controlled models including random site intercepts ($u_{\text{site}}$) for optimal within-UKB accuracy, and (2) transfer models without site effects for application to external cohorts where site is unknown or not comparable. Transfer models used identical distribution families but omitted the random site term, following recommendations for cross-study normative modeling.

### Analysis 5: Brain-Cognition SEM and Measurement Invariance

Joint measurement-structural equation models tested associations between brain measures and cognitive factors. These analyses were restricted to the England subsample (N = `r format(n_england, big.mark=",")`; see Data Source and Participants) because the Index of Multiple Deprivation (IMDP), used as a socioeconomic covariate, is not comparable across UK countries. All models were estimated in lavaan [@rosseel_lavaan_2012].

#### Measurement Invariance

Comparing brain-cognition associations across sexes requires that the cognitive measurement model is equivalent across groups. Measurement invariance of the bifactor structure (see Factor Structure: Bifactor CFA) was therefore tested following standard multi-group CFA procedures [@putnick_measurement_2016]. Four nested models were tested: (1) configural invariance (same factor structure), (2) metric invariance (equal loadings), (3) scalar invariance (equal intercepts), and (4) strict invariance (equal residuals). Invariance was evaluated using $\Delta$CFI $\leq$ 0.01 [@cheung_evaluating_2002] and $\Delta$RMSEA $\leq$ 0.015 [@chen_sensitivity_2007]. At minimum, metric invariance is required to compare structural paths across groups.

#### SEM Specifications

Given adequate measurement invariance, three SEM models were estimated, each combining the bifactor cognitive measurement model with structural paths from a brain measure to all three cognitive factors. Brain measures entered as observed (manifest) variables: HC (bilateral hippocampal volume, sum of left and right), HVR (@eq-hvr), and HC~RES~ (HC residualized against ICV within sex, yielding r $\approx$ 0 with ICV by construction). The structural component of each model took the following form:

**Model 1 (HC):**

$$\begin{aligned}
\text{HC} &= \beta_{\text{Age}} \cdot \text{Age} + \beta_{\text{ICV}} \cdot \text{ICV} + \beta_{\text{Edu}} \cdot \text{Edu} + \beta_{\text{IMDP}} \cdot \text{IMDP} + \zeta_{\text{HC}} \\
\eta_k &= \beta_{k,\text{HC}} \cdot \text{HC} + \beta_{k,\text{Age}} \cdot \text{Age} + \beta_{k,\text{ICV}} \cdot \text{ICV} + \beta_{k,\text{Edu}} \cdot \text{Edu} + \beta_{k,\text{IMDP}} \cdot \text{IMDP} + \zeta_k
\end{aligned}$$ {#eq-sem-hc}

**Model 2 (HVR):**

$$\begin{aligned}
\text{HVR} &= \beta_{\text{Age}} \cdot \text{Age} + \beta_{\text{Age}^2} \cdot \text{Age}^2 + \beta_{\text{ICV}} \cdot \text{ICV} \\
  &\quad + \beta_{\text{Edu}} \cdot \text{Edu} + \beta_{\text{IMDP}} \cdot \text{IMDP} + \zeta_{\text{HVR}} \\
\eta_k &= \beta_{k,\text{HVR}} \cdot \text{HVR} + \beta_{k,\text{Age}} \cdot \text{Age} + \beta_{k,\text{Age}^2} \cdot \text{Age}^2 \\
  &\quad + \beta_{k,\text{ICV}} \cdot \text{ICV} + \beta_{k,\text{Edu}} \cdot \text{Edu} + \beta_{k,\text{IMDP}} \cdot \text{IMDP} + \zeta_k
\end{aligned}$$ {#eq-sem-hvr}

**Model 3 (HC~RES~):**

$$\begin{aligned}
\text{HC}_{\text{RES}} &= \beta_{\text{Age}} \cdot \text{Age} + \beta_{\text{Age}^2} \cdot \text{Age}^2 + \beta_{\text{ICV}} \cdot \text{ICV} \\
  &\quad + \beta_{\text{Edu}} \cdot \text{Edu} + \beta_{\text{IMDP}} \cdot \text{IMDP} + \zeta_{\text{HC}_{\text{RES}}} \\
\eta_k &= \beta_{k,\text{HC}_{\text{RES}}} \cdot \text{HC}_{\text{RES}} + \beta_{k,\text{Age}} \cdot \text{Age} + \beta_{k,\text{Age}^2} \cdot \text{Age}^2 \\
  &\quad + \beta_{k,\text{ICV}} \cdot \text{ICV} + \beta_{k,\text{Edu}} \cdot \text{Edu} + \beta_{k,\text{IMDP}} \cdot \text{IMDP} + \zeta_k
\end{aligned}$$ {#eq-sem-hcres}

where $\eta_k \in \{g, \text{Memory}_s, \text{Speed}_s\}$ denotes latent cognitive factors and $\zeta$ denotes residual terms. A quadratic age term was included in the HVR and HC~RES~ models but not in the HC model, where the combination of raw hippocampal volume, ICV, and quadratic age produced multicollinearity. This difference in model specification means that brain-cognition path comparisons between HC and the other measures should be interpreted with this constraint in mind, as the HC model captures only linear age effects. Age was centered within sex groups before squaring. All structural variables were z-standardized within sex groups to address scale differences. Covariates are summarized in `r S("tbl-sem-covariates-fields")`.

The HC~RES~ model serves as a benchmark representing the methodological standard for removing head-size confounding. Comparing HVR to HC~RES~ tests whether HVR provides similar cognitive associations while being self-normalizing and capturing additional variance through its incorporation of ventricular dynamics. The proportions and stereotaxic adjustment methods were excluded due to concerns about biologically implausible associations and residual ICV correlation, respectively [@wang_comparison_2024].

#### Estimation and Inference

Two estimation approaches were used. Pooled models combined all participants, included sex as a covariate, and were estimated using robust maximum likelihood (MLR) with cluster-robust standard errors to account for scanner site. Multi-group models estimated separate structural parameters for females and males, allowing sex differences in brain-cognition paths to be directly tested; these were estimated using maximum likelihood with bootstrap confidence intervals (2,000 resamples) [@cumming_inference_2005]. Missing cognitive data were handled using full information maximum likelihood (FIML) in both approaches. Sex differences in path coefficients were evaluated by comparing bootstrap confidence intervals across groups.

### Analysis 6: Sensitivity Analyses

Three sensitivity analyses assessed the robustness of sex difference findings to analytical choices.

#### Site effects

Scanner site (UK Biobank assessment centre) can introduce systematic variance in volumetric measurements due to differences in scanner hardware, software versions, or acquisition protocols. Site effects were quantified using the intraclass correlation coefficient:

$$\text{ICC} = \frac{\sigma^2_{\text{site}}}{\sigma^2_{\text{site}} + \sigma^2_{\text{residual}}}$$ {#eq-icc}

Site-adjusted sex difference effect sizes were computed using linear mixed-effects models with a fixed effect for sex and a random intercept for scanner site:

$$Y_i = \beta_0 + \beta_{\text{Sex}} \cdot \text{Sex}_i + u_{\text{site}(i)} + \varepsilon_i$$ {#eq-site-adj}

where $u_{\text{site}} \sim \mathcal{N}(0, \sigma^2_{\text{site}})$ and $\varepsilon_i \sim \mathcal{N}(0, \sigma^2_{\text{residual}})$. Cohen's d was approximated as $\hat{\beta}_{\text{Sex}} / \hat{\sigma}_{\text{residual}}$, with confidence intervals derived from the Wald approximation.

#### Hemisphere comparison

Left and right hippocampus and lateral ventricle volumes (residualized against ICV) were analyzed separately to assess whether sex differences are lateralized. Cohen's d was computed for each hemisphere independently, with effect size comparisons based on overlapping 95% confidence intervals. Non-overlapping intervals indicate a statistically significant difference in effect magnitude between hemispheres.

#### Inclusion of psychiatric diagnoses

The primary analysis excluded participants with ICD-10 diagnoses in the F (mental and behavioral disorders), G (neurological), or Q0 (congenital brain malformations) categories. To assess robustness, sex difference analyses were repeated in an expanded sensitivity sample that included participants with F-code diagnoses (F00-F99) while still excluding G and Q0 codes. The difference in Cohen's d between primary and sensitivity samples ($\Delta d$ = $d_{\text{primary}}$ - $d_{\text{sensitivity}}$) quantified the impact of psychiatric exclusion criteria. Small differences indicate that findings generalize to a less restrictively selected sample and are not driven by exclusion of psychiatric comorbidity.

---

# Results

## Sample Characteristics

The primary sample comprised `r format(n_primary, big.mark=",")` participants (`r format(n_female, big.mark=",")` female, `r format(n_male, big.mark=",")` male; `r pct_female`% female) aged `r age_min`--`r age_max` years (@tbl-demographics). Females were slightly younger on average (M = `r age_mean_f`, SD = `r age_sd_f`) than males (M = `r age_mean_m`, SD = `r age_sd_m`). Males had substantially larger ICV than females, consistent with the well-documented sex difference in head size. The England subsample (N = `r format(n_england, big.mark=",")`), ICV-matched subsample (N = `r format(n_matched_pairs * 2, big.mark=",")`; `r format(n_matched_pairs, big.mark=",")` pairs), and longitudinal subsample (N = 2,935) showed comparable age and ICV distributions to the primary sample within their respective selection constraints.

```{r}
#| label: tbl-demographics
#| tbl-cap: "**Participant characteristics by sex and sample.** Values are M (SD). d = Cohen's d (Female − Male). England = English assessment centres (SEM analyses). Matched = age $\\pm 1$ yr, ICV $\\pm 25$ cc. Longitudinal = participants with repeat imaging."

lng_demog <- if ("LNG" %in% names(demog_data)) demog_data$LNG else NULL
format_demographics_gt(all_demog, eng_demog, mtch_demog, lng_demog)
```

## HVR Self-Normalizing Properties (Analysis 1)

Unadjusted HC showed strong positive correlation with ICV (r = `r round(hvr_icv[VARIABLE == "HC", CORRELATION], 3)`), while proportions-adjusted HC showed moderate negative correlation (r = `r round(hvr_icv[VARIABLE == "HC_PRP", CORRELATION], 3)`), indicating over-correction (@tbl-hvr-validation). Residualization eliminated ICV association by construction (r $\approx$ 0). HVR's correlation with ICV (r = `r round(hvr_icv[VARIABLE == "HVR", CORRELATION], 3)`) was smaller in magnitude than raw HC (r = `r round(hvr_icv[VARIABLE == "HC", CORRELATION], 3)`) but remained negative, unlike residualized HC (r $\approx$ 0). This represents a `r round((1 - abs(hvr_icv[VARIABLE == "HVR", CORRELATION]) / abs(hvr_icv[VARIABLE == "HC", CORRELATION])) * 100, 0)`% reduction in absolute ICV association. The negative direction mirrors the LV-ICV relationship (r = `r round(hvr_icv[VARIABLE == "LV", CORRELATION], 3)`), indicating that the HVR-ICV association reflects the ventricular component's scaling with head size.

```{r}
#| label: tbl-hvr-validation
#| tbl-cap: "**Brain measure correlations with ICV.** An ideal self-normalizing measure shows r(ICV) near zero."

format_hvr_validation_gt(hvr_icv)
```

## Sex Differences in Brain Volumes (Analysis 2)

Sex differences varied across brain measures and adjustment methods (N = `r format(n_primary, big.mark=",")`; @fig-effect-sizes).

### Hippocampal Volume

HC sex differences reversed direction depending on the adjustment method. Unadjusted volumes showed d = `r round(hc_non$ESTIMATE, 2)` [95% CI: `r round(hc_non$CI_LOWER, 2)`, `r round(hc_non$CI_UPPER, 2)`], indicating males had larger absolute volumes. Proportions adjustment reversed this to d = `r round(hc_prp$ESTIMATE, 2)` [`r round(hc_prp$CI_LOWER, 2)`, `r round(hc_prp$CI_UPPER, 2)`] with females larger. Stereotaxic adjustment yielded d = `r round(hc_stx$ESTIMATE, 2)` [`r round(hc_stx$CI_LOWER, 2)`, `r round(hc_stx$CI_UPPER, 2)`], and residualized volumes showed a negligible effect (d = `r round(hc_res$ESTIMATE, 2)` [`r round(hc_res$CI_LOWER, 2)`, `r round(hc_res$CI_UPPER, 2)`]). The effect size range across methods spanned `r round(abs(min(bilateral[ROI == "HC"]$ESTIMATE) - max(bilateral[ROI == "HC"]$ESTIMATE)), 2)` standard deviations.

### Lateral Ventricular Volume

Unlike HC, lateral ventricular sex differences were consistently negative (males > females) across all adjustment methods, though effect magnitudes were attenuated by head-size correction. Unadjusted LV showed d = `r round(lv_non$ESTIMATE, 2)` [`r round(lv_non$CI_LOWER, 2)`, `r round(lv_non$CI_UPPER, 2)`], similar in magnitude to unadjusted HC. After adjustment, all methods retained the male advantage: proportions d = `r round(lv_prp$ESTIMATE, 2)` [`r round(lv_prp$CI_LOWER, 2)`, `r round(lv_prp$CI_UPPER, 2)`], stereotaxic d = `r round(lv_stx$ESTIMATE, 2)` [`r round(lv_stx$CI_LOWER, 2)`, `r round(lv_stx$CI_UPPER, 2)`], and residualized d = `r round(lv_res$ESTIMATE, 2)` [`r round(lv_res$CI_LOWER, 2)`, `r round(lv_res$CI_UPPER, 2)`]. The directional consistency across methods contrasts with the reversals observed for HC.

### HVR

HVR showed a consistent female advantage: d = `r round(hvr_non$ESTIMATE, 2)` [95% CI: `r round(hvr_non$CI_LOWER, 2)`, `r round(hvr_non$CI_UPPER, 2)`]. The combination of males having relatively smaller hippocampi (after head-size adjustment) and larger ventricles produces this female advantage in HVR. Distribution overlap between sexes across adjustment methods is shown in `r S("fig-distributions")`.

```{r}
#| label: fig-effect-sizes
#| fig-cap: "**Sex differences by adjustment method.** Circles = full sample; triangles = ICV-matched sample. Faded points indicate non-significant effects (95% confidence interval crosses zero). Right panel shows $\\Delta d$ = $d_{\\text{full}}$ $-$ $d_{\\text{matched}}$, representing the change in effect size after removing head-size confounding; values near zero indicate head-size-independent effects."
#| fig-width: 6.5
#| fig-height: 5.6
#| dev: cairo_pdf
fig_effect_sizes_by_adjustment(sex_diff, include_matched = TRUE)
```

### Matched-Pair Comparison

Matching on age and ICV (N = `r format(n_matched_pairs * 2, big.mark=",")`; `r format(n_matched_pairs, big.mark=",")` pairs) revealed which sex differences persist after removing head-size confounding (@tbl-matched). HC effects collapsed to near-zero (unadjusted d = `r round(matched_hc_non$ESTIMATE_MTCH, 2)`), confirming that HC sex differences are driven by head-size variation. LV effects were attenuated but remained significant (unadjusted d = `r round(matched_lv_non$ESTIMATE_MTCH, 2)`), indicating that males have larger ventricles even at equivalent head size. HVR was reduced but persisted (d = `r round(matched_hvr$ESTIMATE_MTCH, 2)` [`r round(matched_hvr$CI_LOWER_MTCH, 2)`, `r round(matched_hvr$CI_UPPER_MTCH, 2)`]), consistent with the female HVR advantage being independent of head-size confounding.

```{r}
#| label: tbl-matched
#| tbl-cap: "**Full vs. matched sample effect sizes.** Matching on age ($\\pm 1$ yr) and ICV ($\\pm 25$ cc) eliminates head-size confounding."
if ("COMPARISON" %in% names(sex_diff))
  format_matched_comparison_gt(sex_diff$COMPARISON, n_matched_pairs)
```

## Age x Sex Interactions (Analysis 3)

Cross-sectional age patterns (N = `r format(n_primary, big.mark=",")`) differed by sex (@fig-age-trajectory). For HVR, the age × sex interaction was significant ($\beta$ = `r round(hvr_int$INTERACTION_EST, 4)`, SE = `r round(hvr_int$INTERACTION_SE, 5)`, t = `r round(hvr_int$INTERACTION_T, 2)`, p = `r format(hvr_int$INTERACTION_P, scientific=TRUE, digits=2)`), with males showing `r round(abs(hvr_int$SLOPE_RATIO), 2)`× steeper cross-sectional age-related differences. The female HVR advantage *widened* with age—effect sizes increased from d = `r round(hvr_d_young$ESTIMATE, 2)` at ages 45-55 to d = `r round(hvr_d_old$ESTIMATE, 2)` at ages 75+.

Age × sex interactions were significant for all three measures. HC showed males with `r round(abs(hc_int$SLOPE_RATIO), 2)`× steeper cross-sectional differences (p = `r format(hc_int$INTERACTION_P, scientific=TRUE, digits=2)`). LV showed the largest sex disparity, with males exhibiting `r round(abs(lv_int$SLOPE_RATIO), 2)`× steeper ventricular expansion (p = `r format(lv_int$INTERACTION_P, scientific=TRUE, digits=2)`). The greater sex disparity in ventricular trajectories compared to hippocampal trajectories explains the intermediate HVR ratio (`r round(abs(hvr_int$SLOPE_RATIO), 2)`×), as HVR captures both processes.

```{r}
#| label: fig-age-trajectory
#| fig-cap: "**Age trajectories of sex differences.** Solid line = hippocampal-to-ventricle ratio; dashed lines = hippocampal/ventricular volume. White-filled points = non-significant (95% confidence interval crosses zero)."
#| fig-width: 6.5
#| fig-height: 4.5
fig_age_trajectory(sex_diff)
```

## Normative Modeling (Analysis 4)

@fig-normative-centiles presents sex-specific normative centiles for HVR, hippocampal volume, and lateral ventricular volume. Across the age range, females showed higher HVR values at all percentile levels; the median female HVR at age 55 corresponded approximately to the 75th percentile for males of the same age. HC and LV centile curves are generated from GAMLSS models that include ICV as a covariate, providing head-size adjusted norms without pre-residualizing the data. Males show greater ventricular variability at older ages, potentially reflecting greater heterogeneity in aging trajectories.

```{r}
#| label: fig-normative-centiles
#| fig-cap: "**Sex-specific normative centile curves.** Hippocampal-to-ventricle ratio (top), hippocampal volume (middle), and lateral ventricular volume (bottom). Left and right columns show female and male centiles separately; the centre column overlays both sexes for direct comparison. Hippocampal and ventricular volume models include intracranial volume as a covariate."
#| fig-width: 6.5
#| fig-height: 8
fig_normative_centiles_panel(norm_tables)
```

Calibration plots (`r S("fig-gamlss-calibration")`) demonstrate that normative models accurately capture the distribution of observed values, with observed proportions closely tracking expected proportions across centiles.

### Temporal Stability (Longitudinal Validation)

All brain measures demonstrated high test-retest reliability across the longitudinal follow-up (`r S("tbl-temporal-stability")`), with Pearson correlations > 0.95 and ICCs > 0.93 for all measures. HVR showed slightly lower reliability (r = `r round(hvr_r, 3)`, ICC = `r round(hvr_icc, 3)`) compared to HC (r = `r round(hc_r, 3)`, ICC = `r round(hc_icc, 3)`) and LV (r = `r round(lv_r, 3)`, ICC = `r round(lv_icc, 3)`), consistent with the ratio measure amplifying small measurement errors in either component. These reliabilities support the use of these normative models for individual-level assessment.

### Clinical Normative Centile Tables

Sex-specific normative centile tables for HVR, hippocampal volume, and lateral ventricular volume are provided in Supporting Information (`r S_range("tbl-hvr-centiles-female", "tbl-lv-centiles-male")`) to enable clinical interpretation of individual measurements at 5-year age intervals. These tables allow clinicians to determine an individual's percentile rank given their age, sex, and brain measure.

GAMLSS model coefficients for the transfer models (excluding site random effects) are provided in `r S("tbl-gamlss-hvr-coef")` and `r S("tbl-gamlss-hc-coef")`. These transfer models enable researchers to compute individual centile scores in external datasets where UK Biobank site identifiers are not applicable.

## Brain-Cognition Associations (Analysis 5)

SEM analyses were restricted to the England subsample (N = `r format(n_england, big.mark=",")`) to ensure comparability of the Index of Multiple Deprivation (IMDP) covariate.

### Measurement Invariance

The cognitive bifactor model showed acceptable fit at all invariance levels, with $\Delta$CFI $\leq$ 0.01 and $\Delta$RMSEA $\leq$ 0.015 at each step (`r S("tbl-measurement-invariance")`). Strict measurement invariance was supported, validating cross-sex comparisons of factor scores and structural paths.

### Pooled Model Results

General cognition (*g*) was positively associated with all three brain measures (@tbl-sem-paths): unadjusted HC ($\beta$ = `r safe_round(hc_g$est.std, 3)` [95% CI: `r safe_round(hc_g$ci.lower, 3)`, `r safe_round(hc_g$ci.upper, 3)`]), HVR ($\beta$ = `r safe_round(hvr_g$est.std, 3)` [`r safe_round(hvr_g$ci.lower, 3)`, `r safe_round(hvr_g$ci.upper, 3)`]), and HC~RES~ ($\beta$ = `r safe_round(hc_res_g$est.std, 3)` [`r safe_round(hc_res_g$ci.lower, 3)`, `r safe_round(hc_res_g$ci.upper, 3)`]), all p < .001. Confidence intervals overlapped substantially, indicating no significant differences in predictive validity between measures. HVR showed numerically the strongest association with *g*, despite requiring no head-size adjustment.

For processing speed (Speed~s~), HC showed the strongest effect ($\beta$ = `r safe_round(hc_speed$est.std, 3)`, p < .001), followed by HC~RES~ ($\beta$ = `r safe_round(hc_res_speed$est.std, 3)`, p = `r safe_round(hc_res_speed$pvalue, 3)`), with HVR trending but not significant ($\beta$ = `r safe_round(hvr_speed$est.std, 3)`, p = `r safe_round(hvr_speed$pvalue, 3)`). The memory-specific factor (Memory~s~) could not be reliably estimated in either sex and is therefore not reported here (see `r S("tbl-sem-loadings")` for factor loadings).

```{r}
#| label: tbl-sem-paths
#| tbl-cap: "**Pooled brain-cognition path coefficients.** Beta = standardized path coefficient; CI from 2000 bootstrap resamples. Memory-specific factor paths not shown due to factor collapse."

format_pooled_sem_paths_gt(sem_params, hvr_g, hc_res_g, hc_g)
```

### Sex-Stratified Results

@fig-sem-forest displays sex-stratified path coefficients from multi-group models (`r S("tbl-sem-paths-by-sex")`). The multigroup analyses revealed sex differences in brain-cognition associations.

For general cognition (*g*), HC~RES~ → *g* was not significant in males (p = `r safe_round(hc_res_g_m$pvalue, 3)`) while remaining significant in females. In contrast, HVR showed the opposite pattern: males exhibited stronger HVR-cognition associations ($\beta$ = `r safe_round(hvr_g_m$est.std, 3)`) than females ($\beta$ = `r safe_round(hvr_g_f$est.std, 3)`), with both significant. For processing speed (Speed~s~), males showed significant HC and HC~RES~ associations ($\beta$ = `r safe_round(hc_speed_m$est.std, 3)` and `r safe_round(hc_res_speed_m$est.std, 3)`, respectively) while females showed none, suggesting sex-differential neural substrates for processing speed performance. Neither sex showed significant HVR → Speed~s~ associations.

```{r}
#| label: fig-sem-forest
#| fig-cap: "**Brain-cognition paths by sex.** Points represent standardized coefficients; lines show 95% bootstrap confidence intervals. White-filled points indicate non-significant effects."
#| fig-width: 6.5
#| fig-height: 4.4
fig_sem_forest(sem_params)
```

All SEM models demonstrated good fit according to conventional criteria (`r S("tbl-sem-fit")`; @hu_cutoff_1999). Full model parameters including factor loadings and covariate effects are reported in `r S("tbl-sem-loadings")` and `r S("tbl-sem-covariate-effects")`.

#### Non-Linear Age Effects

As described in Methods, quadratic age terms were included in HVR and HC~RES~ models but not in the HC model due to multicollinearity; fit indices for the HC model (`r S("tbl-sem-fit")`) therefore reflect this linear-age specification. Both brain measures and cognitive factors showed significant negative quadratic age effects (all p < .001; `r S("tbl-sem-covariate-effects")`), indicating steeper cross-sectional differences at older ages. For brain measures, HVR showed $\beta$ = `r safe_round(hvr_age_sq$est.std, 3)` and HC~RES~ showed $\beta$ = `r safe_round(hc_res_age_sq$est.std, 3)`. For cognitive factors, *g* showed $\beta$ = `r safe_round(g_age_sq_hvr$est.std, 3)` in the HVR model and $\beta$ = `r safe_round(g_age_sq_hc$est.std, 3)` in the HC~RES~ model, while Speed~s~ showed $\beta$ = `r safe_round(speed_age_sq_hvr$est.std, 3)` and $\beta$ = `r safe_round(speed_age_sq_hc$est.std, 3)` respectively. The stronger quadratic effect for HVR compared to HC~RES~ suggests that HVR captures both hippocampal atrophy and ventricular expansion, both of which accelerate with age.

## Sensitivity Analyses (Analysis 6)

### Site-Adjusted Effect Sizes

Site ICCs ranged from 0.004 to 0.012 (`r S("tbl-site-adjusted")`), indicating that scanner site explained less than 1.5% of total variance in all brain measures. Effect sizes remained consistent after site adjustment.

### Hemisphere Comparison

Sex difference effect sizes were largely symmetrical across hemispheres for the individual volumes (`r S("tbl-hemisphere")`, `r S("fig-hemisphere")`). Residualized hippocampal volume showed significant effects bilaterally (left d = −0.14, right d = −0.16) with overlapping confidence intervals, indicating no lateralized sex difference. Lateral ventricles similarly showed a male advantage in both hemispheres with nearly identical effect sizes (left d = 0.48, right d = 0.48). However, HVR showed a lateralized pattern: the right hemisphere exhibited a significant female advantage (d = 0.05 [0.03, 0.07]), while the left hemisphere effect was non-significant (d = −0.01 [−0.03, 0.02]).

### Inclusion of Psychiatric Diagnoses (F-Codes)

Effect sizes remained stable when including participants with psychiatric diagnoses (N = 30,032; `r S("tbl-sensitivity")`, `r S("fig-sensitivity")`), demonstrating robustness of findings to exclusion criteria.

---

# Discussion

## Summary of Findings

In `r format(n_primary, big.mark=",")` UK Biobank participants, we evaluated HVR as a self-normalizing biomarker that bypasses conventional head-size correction. HVR showed weak correlation with ICV, substantially reducing head-size dependence compared to unadjusted HC. Hippocampal sex differences ranged from d = `r round(hc_non$ESTIMATE, 2)` (males larger) to d = `r round(hc_prp$ESTIMATE, 2)` (females larger) depending solely on the adjustment method, whereas HVR consistently showed a female advantage replicated in ICV-matched samples. HVR predicted general cognitive ability comparably to hippocampal volume, and males showed steeper age-related HVR differences (`r round(abs(hvr_int$SLOPE_RATIO), 2)`× female slope), with the female advantage widening at older ages (d = `r round(hvr_d_young$ESTIMATE, 2)` at 45-55y to d = `r round(hvr_d_old$ESTIMATE, 2)` at 75+y).

## HVR Self-Normalizing Properties (Analysis 1)

HVR achieves head-size independence through its mathematical construction: both numerator (HC) and denominator (HC + LV) scale with head size, so their ratio largely normalizes out this confound. @schoemaker_hippocampal-ventricle_2019 demonstrated that HVR yielded stronger correlations with both age and memory performance than absolute hippocampal volume, supporting its utility as an integrity index. In the present sample, HVR achieved a `r icv_reduction_pct`% reduction in ICV correlation compared to unadjusted HC (r = `r round(r_hvr_icv, 3)` vs. r = `r round(r_hc_icv, 3)`). This approach complements other ratio measures in clinical neuroimaging, such as the Evans' index [@curra_ventricular_2019], and avoids the pitfalls of proportional ICV correction when brain structures scale non-isometrically with head size [@wang_comparison_2024; @pintzka_marked_2015]. The residual negative HVR-ICV correlation likely reflects ventricular scaling: the hippocampus scales hypoallometrically with ICV while lateral ventricles scale near-isometrically [@de_jong_allometric_2017], so larger heads have disproportionately larger ventricles relative to hippocampal volume, naturally producing lower HVR. Unlike residualization, which removes all ICV association by construction, HVR preserves morphological variation related to how ventricular space scales with brain size.

## Sex Differences in Brain Measures (Analysis 2)

The reversal of HC sex differences across adjustment methods—from d = `r round(hc_non$ESTIMATE, 2)` (males larger, unadjusted) to d = `r round(hc_prp$ESTIMATE, 2)` (females larger, proportions-adjusted), a swing of `r round(abs(hc_d_range), 2)` SD—indicates these differences reflect head-size confounding rather than biology. This over-correction arises because the hippocampus scales hypoallometrically with ICV ($\alpha$ = 0.60; @de_jong_allometric_2017), so dividing by ICV systematically overestimates relative volume in smaller heads [@pintzka_marked_2015; @voevodskaya_effects_2014]. The pattern replicates across multiple cohorts [@pintzka_marked_2015; @voevodskaya_effects_2014; @fjell_minute_2009], and @ritchie_sex_2018 reported that unadjusted hippocampal sex differences in the UK Biobank dropped to near-zero after regression-based adjustment. The near-zero residualized effect (d = `r round(hc_res$ESTIMATE, 3)`) suggests no meaningful sex difference after appropriate statistical control. While total hippocampal volume sex differences appear methodologically determined, sex-specific trajectories in hippocampal subregions [@mu_effect_2020] and structural connectivity patterns [@ingalhalikar_sex_2014] suggest meaningful differences may exist at finer spatial scales. Lateral ventricles, by contrast, showed a consistent male advantage across all methods (@fig-effect-sizes). Because the LV-ICV regression has a negative y-intercept, proportional correction under-corrects rather than over-corrects ventricular volumes—the opposite of what occurs for HC [@pintzka_marked_2015]. The male advantage persists or increases after correction [@fjell_minute_2009], and in our large sample, residualized LV differences remained significant (d = `r round(lv_res$ESTIMATE, 2)`), suggesting that null findings in smaller studies may reflect insufficient power. The consistent male ventricular advantage contrasts sharply with the method-dependent HC reversals, indicating this sex difference is not a head-size artifact.

The matched-pair analysis (@tbl-matched) provides the strongest test of head-size independence. Matching on age and ICV eliminates head-size variation without the mathematical assumptions of statistical correction [@brzezinski-rittner_disentangling_2025]. When matched, the HC sex difference collapsed to d = `r round(matched_hc_non$ESTIMATE_MTCH, 3)` (effectively zero), replicating prior matched-sample findings [@pintzka_marked_2015; @voevodskaya_effects_2014], while LV retained a male advantage (d = `r round(matched_lv_non$ESTIMATE_MTCH, 2)`) and HVR retained a significant female advantage (d = `r round(matched_hvr$ESTIMATE_MTCH, 3)`). The HVR effect decreased `r hvr_reduction_pct`% from full to matched samples, indicating that most of the sex difference is attributable to head-size variation, but a significant residual persists. Since HVR = HC/(HC+LV), the female advantage arises from two converging factors: comparable hippocampal volumes after size matching and relatively smaller female ventricles. This convergence—HC collapsing while LV and HVR effects survive ICV matching—confirms that the female HVR advantage reflects genuine sex differences in the hippocampus-to-ventricle balance, not head-size confounding.

## Age × Sex Interactions (Analysis 3)

Males showed steeper cross-sectional differences across all measures: `r round(hvr_int$SLOPE_RATIO, 2)`× for HVR (p = `r format(hvr_int$INTERACTION_P, scientific=TRUE, digits=2)`), `r round(hc_int$SLOPE_RATIO, 2)`× for HC, and `r round(lv_int$SLOPE_RATIO, 2)`× for LV, consistent with steeper male subcortical decline reported by @ritchie_sex_2018 and steeper male CSF expansion by @fjell_minute_2009. The greater sex disparity in ventricular differences (`r round(lv_int$SLOPE_RATIO, 2)`×) compared to hippocampal differences (`r round(hc_int$SLOPE_RATIO, 2)`×) explains HVR's intermediate ratio: HVR captures both processes, and steeper male ventricular expansion amplifies hippocampal differences. Males also show earlier ventricular abnormality signatures, with ventricular enlargement accelerating in the seventies [@curra_ventricular_2019], consistent with the steeper male HVR differences observed here. However, @brzezinski-rittner_disentangling_2025 found that hippocampal age × sex interactions disappeared in a TIV-matched subsample, suggesting that some sex-differential aging trajectories may be byproducts of initial brain size differences.

The prognostic relevance of trajectory differences is supported by evidence that steeper hippocampal volume loss predicts dementia conversion [@rajagopal_estimating_2024]. Cumulative estrogen exposure may contribute to the female advantage, as longer reproductive spans are associated with younger brain ages [@luders_case_2025] and estrogen promotes hippocampal neurogenesis and synaptic plasticity [@jett_endogenous_2022]. However, the female advantage is not universal: female APOE ε4 carriers show faster hippocampal loss than male carriers in aMCI [@park_sex-specific_2025], indicating that genetic risk can modulate typical sex-differential patterns. Other large studies have found no significant sex differences in hippocampal age slopes [@fjell_minute_2009; @pintzka_marked_2015], underscoring that the presence of age × sex interactions depends on sample size, age range, and head-size adjustment method. These cross-sectional patterns cannot establish true individual trajectories and may be confounded by cohort effects or differential survival; longitudinal validation is essential.

## Normative Modeling (Analysis 4)

The sex-specific centile curves (@fig-normative-centiles) extend normative brain charts [@bethlehem_brain_2022; @dima_subcortical_2022; @ge_normative_2024] by providing the first sex-specific reference values for HVR. Our GAMLSS approach aligns with @bethlehem_brain_2022 and @dima_subcortical_2022, flexibly modeling both non-linear age trends and age-related variability changes. Sex-specific norms are essential to avoid misclassification, consistent with recommendations from @williams_neuroanatomical_2021 and @ge_normative_2024, who found that sex accounts for a considerable proportion of morphological variance. For residualized HC, sex differences in centiles are minimal, consistent with the near-zero Cohen's d for residualized volumes, supporting combined norms when proper ICV adjustment is applied. For LV, males show greater ventricular variability at older ages, potentially reflecting greater heterogeneity in aging-related patterns. Well-calibrated centile estimates (`r S("fig-gamlss-calibration")`) have demonstrated clinical utility in detecting case-control differences and predicting diagnostic transition from MCI to Alzheimer's disease [@bethlehem_brain_2022; @dima_subcortical_2022].

## Brain-Cognition Associations (Analysis 5)

Strict measurement invariance for the cognitive bifactor model ($\Delta$CFI $\leq$ 0.01 and $\Delta$RMSEA $\leq$ 0.015 at each level; `r S("tbl-measurement-invariance")`) confirmed that the same cognitive constructs (*g*, Memory~s~, Speed~s~) exist in both sexes with equivalent loadings, intercepts, and residual variances [@chen_sensitivity_2007; @putnick_measurement_2016], validating cross-sex comparison of brain-cognition paths. In pooled models, HVR predicted general cognition comparably to HC without requiring head-size adjustment (HVR → *g* $\beta$ = `r safe_round(hvr_g$est.std, 3)` vs. HC → *g* $\beta$ = `r safe_round(hc_g$est.std, 3)`, overlapping CIs), with all effects small—consistent with the modest brain-cognition associations typically observed in large healthy samples [@ritchie_sex_2018; @wang_comparison_2024]. The restricted cognitive range in healthy volunteer samples likely attenuates these associations [@bachmann_age-_2023]. Multigroup analyses revealed sex-differential patterns: HC~RES~ predicted cognition in females but not males ($\beta$ = `r safe_round(hc_res_g_m$est.std, 3)`, p = `r safe_round(hc_res_g_m$pvalue, 3)`), consistent with stronger female hippocampal-cognition associations reported elsewhere [@zheng_sex_2017; @jockwitz_differential_2024]. Raw HC showed significant associations in both sexes but with a female advantage ($\beta$ = `r safe_round(hc_g_f$est.std, 3)` vs. `r safe_round(hc_g_m$est.std, 3)`). HVR showed the opposite pattern, with males exhibiting stronger associations ($\beta$ = `r safe_round(hvr_g_m$est.std, 3)` vs. `r safe_round(hvr_g_f$est.std, 3)`), though both sexes reached significance.

These sex-differential patterns indicate that HVR is not simply "residualized HC by another means." Unlike HC~RES~, which lost predictive validity in males, HVR maintained significant associations in both sexes—paralleling @schoemaker_hippocampal-ventricle_2019 and @fernandez-lozano_enhanced_2025, who found that HVR predicted cognition comparably or better than HC alone. The differential patterns between HC~RES~ and HVR across sexes suggest HVR captures distinct information related to hippocampus-ventricle coupling. HVR also accommodated quadratic age specifications that raw HC could not (the HC + ICV + age² combination produced multicollinearity), allowing detection of accelerating age-related patterns ($\beta$~age²~ = `r safe_round(hvr_age_sq$est.std, 3)` for HVR vs. `r safe_round(hc_res_age_sq$est.std, 3)` for HC~RES~). The stronger quadratic effect for HVR reflects its capture of both hippocampal and ventricular age-related differences. The ventricular component likely contributes cognitive-relevant variance beyond medial temporal integrity, as ventricular volume has been associated with executive function rather than episodic memory [@bachmann_age-_2023].

## Sensitivity Analyses (Analysis 6)

Low site ICCs (all < 0.05; `r S("tbl-site-adjusted")`) confirmed minimal scanner site contribution to variance, and effect sizes were stable after site adjustment; the ratio construction of HVR did not amplify site-related noise. Sex difference effect sizes were symmetrical across hemispheres for HC and LV; HVR showed a small lateralized pattern (right d = 0.05 vs. left d = −0.01), though the small magnitude and potential contribution of hemispheric segmentation biases [@fernandez-lozano_enhanced_2025] suggest bilateral HVR remains most robust for clinical application. Including participants with psychiatric diagnoses changed effect sizes minimally ($\Delta$d < 0.02; `r S("tbl-sensitivity")`, `r S("fig-sensitivity")`), consistent with the low prevalence of psychiatric diagnoses in this healthy volunteer cohort [@fry_comparison_2017].

## Limitations

Design and sample constraints include the cross-sectional design, which cannot distinguish age effects from cohort effects or establish individual change trajectories, and HVR's conflation of hippocampal and ventricular contributions to observed differences. UK Biobank participants are healthier and more educated than the general population [@fry_comparison_2017], limiting generalizability to clinical populations. Some cognitive tests had incomplete data, and complete-case analyses may introduce selection bias. APOE genotype was unavailable, precluding examination of gene × sex interactions relevant to dementia risk.

Methodologically, the study did not include power-corrected proportions allometric adjustment [@sanchis-segura_effects_2020]. Imaging site clustering was not modeled in multigroup SEM because lavaan does not support this combination [@rosseel_lavaan_2012], though the minimal design effect across 4 sites makes sex comparisons conservative rather than biased. The bifactor model's memory-specific factor could not be reliably estimated in either sex—a common occurrence in bifactor models [@eid_anomalous_2017; @burns_application_2020]—indicating that memory variance was fully captured by *g*; brain-cognition paths involving Memory~s~ are therefore not reported. Ratios can exhibit non-normality and complex interpretation when numerator and denominator correlate differently with outcomes, though HVR showed approximately normal distributions here. All brain-cognition associations were small ($\beta$ = 0.02–0.04), explaining less than 1% of variance in cognitive performance. While statistically significant given the large sample, the practical and clinical significance of these associations remains to be established.

## Future Directions

The current findings motivate four lines of follow-up work. First, the UK Biobank repeat imaging subsample enables longitudinal validation of whether HVR predicts individual-level cognitive decline. Second, clinical validation in AD/MCI cohorts would assess HVR's diagnostic utility compared to clinical populations studied previously [@fernandez-lozano_enhanced_2025]. Third, longitudinal data could separate hippocampal and ventricular contributions to HVR changes, which the cross-sectional ratio conflates. Fourth, examining whether HVR × sex × APOE genotype predicts dementia conversion would address the gene × sex interactions that this study could not explore.

## Conclusions

In `r format(n_primary, big.mark=",")` UK Biobank participants (`r pct_female`% female), HVR demonstrated substantial head-size independence (r~ICV~ = `r round(r_hvr_icv, 2)`; `r icv_reduction_pct`% reduction vs. raw HC), a consistent female advantage replicated in ICV-matched samples (d = `r round(matched_hvr$ESTIMATE_MTCH, 2)`), and comparable brain-cognition associations to hippocampal volume (HVR → *g*: $\beta$ = `r safe_round(hvr_g$est.std, 3)` vs. HC → *g*: $\beta$ = `r safe_round(hc_g$est.std, 3)`, overlapping CIs). Hippocampal sex differences reversed direction across adjustment methods (d range: `r round(abs(hc_d_range), 1)` SD), confirming their methodological determination. HVR maintained predictive validity in both sexes, while residualized HC lost significance in males (p = `r safe_round(hc_res_g_m$pvalue, 3)`), and supported more complex SEM specifications that raw HC could not. Males showed `r round(hvr_int$SLOPE_RATIO, 2)`× steeper age-related HVR differences (p = `r format(hvr_int$INTERACTION_P, scientific=TRUE, digits=2)`). Sex-specific normative centile curves are provided for research application, pending validation in clinical populations.

# Author Contributions {.unnumbered}

S.F-L. conceived the study, developed the analysis pipeline, performed all analyses, and wrote the manuscript. D.L.C. supervised the project, provided critical review, and approved the final manuscript.

# Acknowledgments {.unnumbered}

This research has been conducted using the UK Biobank Resource under Application Number 45551. This investigation was supported in part by an award from the International Progressive MS Alliance (award reference number PA-1412-02420), a grant from the Canadian Institutes of Health Research, and a donation from the Famille Louise & Andre Charron.

# Conflict of Interest Statement {.unnumbered}

The authors declare no conflicts of interest.

# Data Availability Statement {.unnumbered}

This research was conducted using UK Biobank data (Application 45551). Data are available to approved researchers through the UK Biobank Access Management System (https://www.ukbiobank.ac.uk/). Analysis code is available at https://github.com/soffiafdz/hvr-sex-differences.

# Ethics Approval Statement {.unnumbered}

UK Biobank received ethical approval from the North West Multi-centre Research Ethics Committee (REC reference: 16/NW/0274). All participants provided written informed consent.

# References {.unnumbered}

::: {#refs}
:::
